---
title: "Data Sources & Quality Assurance"
subtitle: "SMARD · ENTSO-E · Copernicus ERA5 — hourly, 2020–2025"
description: |
  A structured inventory of the three primary data sources used throughout 
  this analysis, including coverage checks, distributional profiling, and 
  a documented treatment of the one material irregularity: reporting-window 
  gaps in ENTSO-E cross-border physical flows.
---

```{r}
#| label: r-setup
#| message: false
#| warning: false
#| include: false

knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE
)

library(tidyverse)
library(arrow)
library(scales)
library(patchwork)

theme_set(
  theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold", size = 13),
      plot.subtitle = element_text(colour = "grey40", size = 10),
      plot.caption = element_text(colour = "grey50", size = 8),
      strip.text = element_text(face = "bold"),
      panel.grid.minor = element_blank()
    )
)
```

## Overview

This page inventories every dataset feeding the analysis pipeline and documents
their temporal coverage, resolution, and quality. The objective is twofold:
give the reader confidence that the raw materials are sound, and establish a
clean, aligned hourly time index that all downstream stages can join on without
surprises.

Three sources contribute:

| Source | Domain | Resolution | Period | Access |
|--------|--------|------------|--------|--------|
| **SMARD** (BNetzA) | DE-LU prices, generation by fuel type, load | 15 min / hourly | 2020–2025 | REST API (CC BY 4.0) |
| **ENTSO-E** Transparency | Day-ahead prices (multi-zone), cross-border flows, generation | Hourly | 2020–2025 | `entsoe-py` → Parquet |
| **Copernicus ERA5** | 100 m wind speed, surface solar irradiance, 2 m temperature | Hourly | 2020–2025 | `cdsapi` → NetCDF → Parquet |

All timestamps are stored in UTC. Timezone-aware joins (particularly across
CET/CEST boundaries) are handled at the point of analysis, not in the raw
files.


## 1.0 SMARD (DE-LU) {#sec-smard}

SMARD is the Bundesnetzagentur's public market-data portal for Germany–
Luxembourg. It provides the highest-resolution view of the German power
system: generation by source, total load, and day-ahead prices at 15-minute
granularity.

```{r}
#| label: smard-data
#| code-summary: "SMARD ingestion & profile"

smard <- read_parquet("data/processed/smard/smard_DE-LU_2020_2025.parquet")

tibble(
  Metric = c("Rows", "Columns", "First timestamp (UTC)", 
             "Last timestamp (UTC)", "Total NAs"),
  Value  = c(
    format(nrow(smard), big.mark = ","),
    ncol(smard),
    as.character(min(smard$datetime_utc)),
    as.character(max(smard$datetime_utc)),
    sum(is.na(smard))
  )
) |> knitr::kable()
```

```{r}
#| label: smard-na-counts
#| code-summary: "Column-level NA counts"

enframe(colSums(is.na(smard)), name = "Column", value = "NAs") |>
  knitr::kable()
```

### 1.1 Distributions {#sec-smard-dists}

```{r}
#| label: fig-smard-distributions
#| fig-cap: "Marginal distributions of all SMARD variables (DE-LU, 2020–2025, hourly)."
#| fig-width: 10
#| fig-height: 8
#| code-summary: "Histogram facets"

smard |>
  select(-datetime_utc) |>
  pivot_longer(everything(), names_to = "variable", values_to = "value") |>
  ggplot(aes(x = value)) +
  geom_histogram(bins = 80, fill = "steelblue", alpha = 0.7) +
  facet_wrap(~ variable, scales = "free", ncol = 3) +
  labs(
    title = "SMARD (DE-LU) variable distributions",
    subtitle = "2020–2025, hourly resolution",
    x = NULL, y = "Count",
    caption = "Source: SMARD / Bundesnetzagentur (CC BY 4.0)"
  )
```

**Takeaway:** The SMARD dataset is complete with no missing values across the
full 2020–2025 window. Price distributions exhibit the expected heavy right
tail (scarcity events) and occasional negative prices (renewable surplus hours).
Generation variables show the characteristic bimodal structure of wind output
and the zero-inflated daytime peak of solar.


## 2.0 ERA5 Reanalysis (DE-LU zone) {#sec-era5}

Copernicus ERA5 provides the meteorological backbone of the analysis: the
physical weather drivers that determine how much renewable generation the
installed fleet can produce in any given hour. Variables were downloaded at
native 0.25° resolution and aggregated to a DE-LU zone average using
cosine-latitude area weighting.

```{r}
#| label: era5-data
#| code-summary: "ERA5 ingestion & profile"

era5 <- read_parquet("data/processed/era5/DE-LU.parquet")

tibble(
  Metric = c("Rows", "Columns", "First timestamp (UTC)", 
             "Last timestamp (UTC)", "Total NAs"),
  Value  = c(
    format(nrow(era5), big.mark = ","),
    ncol(era5),
    as.character(min(era5$datetime_utc)),
    as.character(max(era5$datetime_utc)),
    sum(is.na(era5))
  )
) |> knitr::kable()
```

### 2.1 Temporal continuity check {#sec-era5-gaps}

A critical assumption for all downstream modelling is that the ERA5 series has
no gaps — every hour in the analysis window must be present exactly once.

```{r}
#| label: era5-gap-check
#| code-summary: "Gap detection"

era5_ts <- era5 |>
  arrange(datetime_utc) |>
  mutate(gap_hours = as.numeric(difftime(datetime_utc, lag(datetime_utc), units = "hours")))

gap_tab <- table(era5_ts$gap_hours)
cat("Hour-gap distribution:\n")
print(gap_tab)

if (all(names(gap_tab) %in% c("NA", "1"))) {
  cat("\n✓ No gaps detected — series is contiguous at hourly resolution.\n")
} else {
  cat("\n⚠ Non-unit gaps found — investigate before proceeding.\n")
}
```

### 2.2 Distributions {#sec-era5-dists}

```{r}
#| label: fig-era5-distributions
#| fig-cap: "ERA5 variable distributions (DE-LU, 2020–2025). Solar irradiance filtered to daytime hours (> 0 W/m²) to reveal the non-trivial part of the distribution."
#| fig-width: 9
#| fig-height: 7
#| code-summary: "ERA5 histograms"

p_all <- era5 |>
  select(-datetime_utc) |>
  pivot_longer(everything(), names_to = "variable", values_to = "value") |>
  ggplot(aes(x = value)) +
  geom_histogram(bins = 80, fill = "steelblue", alpha = 0.7) +
  facet_wrap(~ variable, scales = "free", ncol = 2) +
  labs(
    title = "ERA5 (DE-LU) — all hours",
    x = NULL, y = "Count"
  )

p_daytime <- era5 |>
  select(-datetime_utc) |>
  pivot_longer(everything(), names_to = "variable", values_to = "value") |>
  mutate(
    variable = case_when(
      variable == "ssrd_wm2" & value > 0 ~ "ssrd_wm2 (daytime only)",
      TRUE ~ variable
    )
  ) |>
  filter(!(variable == "ssrd_wm2" & value == 0)) |>
  ggplot(aes(x = value)) +
  geom_histogram(bins = 80, fill = "steelblue", alpha = 0.7) +
  facet_wrap(~ variable, scales = "free", ncol = 2) +
  labs(
    title = "ERA5 (DE-LU) — SSRD filtered to daytime",
    x = NULL, y = "Count",
    caption = "Source: Copernicus ERA5 reanalysis (Hersbach et al., 2020)"
  )

p_all / p_daytime
```

**Takeaway:** ERA5 is complete and contiguous. Wind speed distributions are
right-skewed (as expected for a Weibull-distributed variable), and 2 m
temperature shows the familiar bimodal winter/summer structure of a
continental-temperate climate. Surface solar irradiance is zero-inflated by
nighttime hours; the daytime-filtered panel reveals the unimodal distribution
that will drive the capacity factor model.


## 3.0 ENTSO-E Transparency Platform {#sec-entsoe}

ENTSO-E supplies three datasets used in this project: day-ahead prices across
multiple bidding zones, cross-border physical flows, and actual generation by
production type. All were fetched via `entsoe-py` and stored as Parquet.

### 3.1 Day-ahead prices {#sec-entsoe-prices}

```{r}
#| label: entsoe-prices-data
#| code-summary: "ENTSO-E price profile"

prices <- read_parquet("data/processed/entsoe/prices_2020_2025.parquet")

tibble(
  Metric = c("Rows", "Columns (zones + timestamp)", 
             "First timestamp (UTC)", "Last timestamp (UTC)", "Total NAs"),
  Value  = c(
    format(nrow(prices), big.mark = ","),
    ncol(prices),
    as.character(min(prices$datetime_utc)),
    as.character(max(prices$datetime_utc)),
    sum(is.na(prices))
  )
) |> knitr::kable()
```

```{r}
#| label: entsoe-prices-summary
#| code-summary: "Price summary statistics (EUR/MWh)"

prices |>
  select(-datetime_utc) |>
  summary() |> t() |>
  knitr::kable()
```


### 3.2 Cross-border physical flows {#sec-flows}

Cross-border flows are the one dataset with a material data-quality issue.
Several flow pairs — particularly those involving Denmark (DK_1, DK_2) and
France — exhibit extended periods of complete missingness.

```{r}
#| label: entsoe-flows-data
#| code-summary: "Flow dataset profile"

flows <- read_parquet("data/processed/entsoe/crossborder_flows_2020_2025.parquet")

tibble(
  Metric = c("Rows", "Flow pairs", "First timestamp (UTC)", 
             "Last timestamp (UTC)"),
  Value  = c(
    format(nrow(flows), big.mark = ","),
    ncol(flows) - 1,
    as.character(min(flows$datetime_utc)),
    as.character(max(flows$datetime_utc))
  )
) |> knitr::kable()
```

```{r}
#| label: entsoe-flows-na-counts
#| code-summary: "Non-zero NA counts by flow pair"

na_counts <- colSums(is.na(flows))
na_nonzero <- na_counts[na_counts > 0]

if (length(na_nonzero) > 0) {
  enframe(na_nonzero, name = "Flow pair", value = "Missing hours") |>
    mutate(
      `% missing` = percent(
        `Missing hours` / nrow(flows), accuracy = 0.1
      )
    ) |>
    arrange(desc(`Missing hours`)) |>
    knitr::kable()
} else {
  cat("No missing values found.\n")
}
```

#### 3.2.1 Diagnosing the gap pattern {#sec-flows-diagnosis}

The heatmap below shows the monthly NA rate for every flow pair. The pattern is
consistent with a **change in ENTSO-E reporting granularity or data provider
switchover** rather than random data loss: gaps begin (or end) cleanly at
month boundaries and affect entire flow pairs simultaneously.

```{r}
#| label: fig-flows-na-heatmap
#| fig-cap: "Monthly NA rate by cross-border flow pair. Blue = complete, red = missing. Gaps align with calendar boundaries, suggesting a reporting-window change rather than random data loss."
#| fig-width: 10
#| fig-height: 7
#| code-summary: "Flow NA heatmap"

flows |>
  mutate(month = lubridate::floor_date(datetime_utc, "month")) |>
  select(-datetime_utc) |>
  group_by(month) |>
  summarise(across(everything(), ~ mean(is.na(.x))), .groups = "drop") |>
  pivot_longer(-month, names_to = "flow_pair", values_to = "pct_missing") |>
  ggplot(aes(x = month, y = flow_pair, fill = pct_missing)) +
  geom_tile() +
  scale_fill_gradient2(
    low = "steelblue", mid = "white", high = "firebrick",
    midpoint = 0.5, limits = c(0, 1),
    labels = percent,
    name = "% Missing"
  ) +
  scale_x_datetime(date_breaks = "6 months", date_labels = "%b\n%Y") +
  labs(
    title = "ENTSO-E cross-border flow data availability",
    subtitle = "Monthly NA rate by flow pair",
    x = NULL, y = NULL,
    caption = "Source: ENTSO-E Transparency Platform via entsoe-py"
  ) +
  theme(
    axis.text.y = element_text(size = 7),
    panel.grid = element_blank()
  )
```



#### 3.2.2 Treatment strategy {#sec-flows-treatment}

The gaps affect a subset of flow pairs (primarily DK and FR interconnectors)
and are concentrated in specific calendar windows. For downstream modelling
in Stages 2–3, we adopt the following approach:

1. **Core flow pairs** (AT, CH, CZ, NL, PL ↔ DE-LU) are complete or
   near-complete across the full window and will be used without imputation.
2. **Partially available pairs** (DK, FR, Nordic interconnectors) will be
   included only in time windows where they are available. The net export
   feature constructed in Stage 2 will sum over whichever pairs are non-missing
   at each timestamp.
3. **No imputation** is applied to flow data. Imputing cross-border flows
   would inject false precision into a variable that is itself a market outcome
   — it is preferable to let the regression handle partial coverage via the
   summed net-export feature.

This is a conservative approach. It means the net export variable will
undercount total physical flows during gap periods, which may attenuate its
estimated coefficient in the price regression. We flag this as a known
limitation in the Stage 2 discussion.


### 3.3 Actual generation by production type {#sec-entsoe-generation}

```{r}
#| label: entsoe-generation-inventory
#| code-summary: "Generation file inventory"

gen_files <- list.files("data/processed/entsoe", 
                        pattern = "generation_", full.names = TRUE)

map_dfr(gen_files, function(f) {
  df <- read_parquet(f)
  tibble(
    File   = basename(f),
    Rows   = format(nrow(df), big.mark = ","),
    Cols   = ncol(df),
    From   = as.character(min(df$datetime_utc)),
    To     = as.character(max(df$datetime_utc)),
    NAs    = sum(is.na(df))
  )
}) |> knitr::kable()
```

::: {.callout-note title="Nuclear generation and Germany's Atomausstieg"}
If nuclear generation columns contain NAs or zeros from mid-2023 onward, this
is not a data-quality issue — it reflects the completion of Germany's nuclear
phase-out (*Atomausstieg*). Three reactors (Brokdorf, Grohnde, and
Gundremmingen C) were permanently shut down on 31 December 2021, halving the
remaining fleet. The final three plants — Emsland, Isar 2, and
Neckarwestheim 2 — operated in fuel stretch-out mode (no new fuel elements
permitted) until they were disconnected from the grid on 15 April 2023,
ending over six decades of commercial nuclear power in Germany.
([BASE — Federal Office for the Safety of Nuclear Waste Management](https://www.base.bund.de/en/nuclear-safety/nuclear-phase-out/nuclear-phase-out_content.html))

In the context of this analysis, the nuclear shutdown is analytically
significant: it removed approximately 4 GW of baseload capacity from the
DE-LU merit order, increasing the residual load that must be met by
dispatchable thermal and imported power — and thereby amplifying the
price sensitivity to renewable variability that Stage 2 aims to quantify.
:::


## 4.0 Alignment & readiness {#sec-alignment}

All three sources share a common UTC hourly index across 2020–2025. The table
below summarises alignment status:

| Check | Status |
|-------|--------|
| Common temporal resolution | ✓ Hourly (SMARD aggregated from 15 min) |
| Common timezone | ✓ All stored as UTC |
| Date range overlap | ✓ 2020-01-01 through 2025 (varies by source) |
| Missing data | ✓ None in SMARD or ERA5; documented gaps in ENTSO-E flows (see @sec-flows) |
| Join key | `datetime_utc` across all tables |

The data are ready for Stage 1 (Weather → Generation modelling) without
further cleaning. The cross-border flow gaps are the only irregularity and
are handled by construction in the net-export feature (Stage 2).
