---
title: "Weather → Generation"
subtitle: "Cross-zone weather profiles and DE-LU capacity factor models"
params:
  analysis_year_start: 2020
  analysis_year_end: 2025
format:
  html:
    code-fold: true
    code-summary: "Show code"
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 9,
  fig.height = 5
)

library(tidyverse)
library(tidymodels)
library(tsibble)
library(arrow)
library(patchwork)
library(lubridate)
library(glue)

# ── DE-LU is the modelling zone ─────────────────────────────────────
zone_file <- "DE-LU"

# ── Custom theme (consistent with 01_data.qmd) ──────────────────────
theme_report <- theme_minimal(base_family = "Arial", base_size = 12) +
  theme(
    plot.title    = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(colour = "grey40", size = 11),
    plot.caption  = element_text(colour = "grey50", size = 8),
    strip.text    = element_text(face = "bold"),
    legend.position = "bottom"
  )

theme_set(theme_report)

# ── Colour palette ───────────────────────────────────────────────────
pal <- c(
  wind_onshore  = "#2166AC",
  wind_offshore = "#67A9CF",
  solar         = "#F4A582"
)
```


## 1.0 Overview {#sec-overview}

This page has two parts. First, we compare **ERA5 reanalysis weather
profiles across all available bidding zones** — wind speed distributions,
solar irradiance gradients, and seasonal patterns — to build intuition
for why renewable generation (and its price impact) differs across
Central Western Europe and the Nordics.

Second, we **fit capacity-factor models for DE-LU**, where SMARD
provides the most granular generation breakdown by technology. These
models link ERA5 weather variables to observed wind onshore, wind
offshore, and solar PV output, and feed directly into the price
decomposition (Stage 2) and risk simulation (Stage 3).

**Modelling approach.** For each technology we:

1. Join hourly ERA5 weather to observed SMARD generation and compute
   capacity factors using BNetzA installed-capacity milestones.
2. Fit a `tidymodels` workflow — polynomial weather terms plus
   month/hour fixed effects (with month × irradiance interactions for
   solar) — on a temporal 80/20 train/test split.
3. Evaluate with RMSE, MAE, and R² on the held-out test period.


## 2.0 Data loading {#sec-data}

```{r}
#| label: load-data

# ── Load all available ERA5 zone parquets ────────────────────────────
era5_files <- list.files("data/processed/era5",
                         pattern = "\\.parquet$",
                         full.names = TRUE)

era5_all <- map(era5_files, \(f) {
  read_parquet(f) |>
    mutate(zone = str_remove(basename(f), "\\.parquet$"))
}) |>
  list_rbind()

zones_available <- sort(unique(era5_all$zone))
cat("ERA5 zones loaded:", paste(zones_available, collapse = ", "), "\n")
cat("Total ERA5 rows:", nrow(era5_all), "\n")

# ── Load DE-LU ERA5 and SMARD separately for modelling ──────────────
era5 <- era5_all |> filter(zone == zone_file) |> select(-zone)

smard_path <- list.files("data/processed/smard",
                         pattern = glue("smard_{zone_file}"),
                         full.names = TRUE)
smard <- read_parquet(smard_path)

cat("\nDE-LU ERA5 rows:", nrow(era5),
    " | range:", format(min(era5$datetime_utc)), "–",
    format(max(era5$datetime_utc)), "\n")
cat("DE-LU SMARD rows:", nrow(smard),
    " | range:", format(min(smard$datetime_utc)), "–",
    format(max(smard$datetime_utc)), "\n")
```


## 2.1 Derived weather features: HDD & CDD {#sec-hdd-cdd}

Heating degree days (HDD) and cooling degree days (CDD) translate raw
temperature into demand-relevant signals. When temperatures drop below
the base (18 °C), heating load rises roughly in proportion; when they
exceed it, cooling demand kicks in. These are standard features in
electricity load and price modelling across European markets.

We compute hourly HDD and CDD here and save them alongside the existing
ERA5 variables so that Stage 2 (price decomposition) can ingest them
directly.

```{r}
#| label: derive-hdd-cdd

hdd_cdd_base <- 18  # °C — standard European base temperature

era5_all <- era5_all |>
  mutate(
    hdd = pmax(hdd_cdd_base - temperature_2m, 0),
    cdd = pmax(temperature_2m - hdd_cdd_base, 0)
  )

# Update DE-LU subset
era5 <- era5_all |> filter(zone == zone_file) |> select(-zone)

# Summary check
era5 |>
  summarise(
    `HDD mean` = round(mean(hdd), 2),
    `HDD max`  = round(max(hdd), 1),
    `CDD mean` = round(mean(cdd), 2),
    `CDD max`  = round(max(cdd), 1),
    `Hours T < 18°C` = sum(temperature_2m < hdd_cdd_base),
    `Hours T > 18°C` = sum(temperature_2m > hdd_cdd_base)
  ) |>
  knitr::kable(caption = "HDD/CDD summary (DE-LU, 18 °C base)")
```

```{r}
#| label: fig-hdd-cdd-seasonal
#| fig-cap: "Seasonal distribution of hourly HDD and CDD (DE-LU). HDD dominates in winter; CDD is concentrated in summer but much smaller in magnitude — consistent with Germany's heating-dominated climate."
#| fig-width: 10
#| fig-height: 5

era5 |>
  mutate(
    season = case_when(
      month(datetime_utc) %in% c(12, 1, 2) ~ "Winter",
      month(datetime_utc) %in% c(3, 4, 5)  ~ "Spring",
      month(datetime_utc) %in% c(6, 7, 8)  ~ "Summer",
      TRUE                                  ~ "Autumn"
    ) |> factor(levels = c("Winter", "Spring", "Summer", "Autumn"))
  ) |>
  pivot_longer(c(hdd, cdd), names_to = "variable", values_to = "value") |>
  filter(value > 0) |>
  mutate(variable = recode(variable, hdd = "HDD (18°C base)", cdd = "CDD (18°C base)")) |>
  ggplot(aes(x = value, fill = season)) +
  geom_histogram(bins = 50, alpha = 0.7, position = "identity") +
  facet_wrap(~ variable, scales = "free") +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title   = "Heating & Cooling Degree Distribution by Season",
    subtitle = "Hourly values > 0, DE-LU (18 °C base temperature)",
    x = "Degree hours", y = "Count", fill = NULL,
    caption = "Source: Copernicus ERA5 reanalysis"
  )
```


## 3.0 Cross-zone weather profiles {#sec-cross-zone}

Before narrowing to DE-LU for capacity-factor modelling, we compare the
weather resource across all zones for which ERA5 spatial averages are
available. These profiles explain much of the cross-zone variation in
renewable generation — and ultimately in price dynamics — explored in
Stage 2.


### 3.1 Wind speed distributions {#sec-cross-zone-wind}

Hub-height (100 m) wind speed distributions differ markedly across
bidding zones, reflecting geographic exposure to Atlantic weather
systems, coastal effects, and orography. Zones with higher mean wind
speeds and heavier right tails support higher average capacity factors
but also contribute more generation variability.

```{r}
library(plotly)

wind_density <- era5_all |>
  group_by(zone) |>
  group_map(\(df, grp) {
    d <- density(df$wind_speed_100m, from = 0, n = 256)
    tibble(zone = grp$zone, wind_speed = d$x, density = d$y)
  }) |>
  list_rbind()

plot_ly(wind_density, x = ~wind_speed, y = ~density,
        color = ~zone, type = "scatter", mode = "lines",
        fill = "tozeroy", opacity = 0.35,
        hovertemplate = paste0(
          "<b>%{meta}</b><br>",
          "Wind speed: %{x:.1f} m/s<br>",
          "Density: %{y:.4f}<extra></extra>"
        ),
        meta = ~zone) |>
  layout(
    title = list(
      text = "100m wind speed distributions by bidding zone",
      font = list(size = 14)
    ),
    xaxis = list(title = "Wind speed (m/s)"),
    yaxis = list(title = "Density"),
    legend = list(title = list(text = "Zone")),
    annotations = list(
      list(text = "Source: Copernicus ERA5 reanalysis",
           xref = "paper", yref = "paper", x = 1, y = -0.12,
           showarrow = FALSE, font = list(size = 8, color = "grey50"),
           xanchor = "right")
    )
  )
```



### 3.2 Seasonal wind profiles {#sec-cross-zone-seasonal-wind}

Wind resource shows strong seasonality across all zones — winter months
are consistently windier — but the magnitude of the seasonal swing
varies geographically. Zones with a large winter-to-summer ratio face
more pronounced seasonal generation risk.

```{r}
#| label: fig-cross-zone-seasonal-wind
#| fig-height: 6
#| fig-width: 10

era5_all |>
  mutate(
    season = case_when(
      month(datetime_utc) %in% c(12, 1, 2) ~ "Winter",
      month(datetime_utc) %in% c(3, 4, 5)  ~ "Spring",
      month(datetime_utc) %in% c(6, 7, 8)  ~ "Summer",
      TRUE                                  ~ "Autumn"
    ),
    season = factor(season, levels = c("Winter", "Spring", "Summer", "Autumn"))
  ) |>
  group_by(zone, season) |>
  summarise(
    mean_wind = mean(wind_speed_100m, na.rm = TRUE),
    .groups   = "drop"
  ) |>
  mutate(zone = fct_reorder(zone, mean_wind, .fun = mean, .desc = TRUE)) |>
  ggplot(aes(x = zone, y = mean_wind, fill = season)) +
  geom_col(position = "dodge", alpha = 0.85) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title    = "Mean 100 m wind speed by zone and season",
    subtitle = "Winter dominance is universal but the seasonal swing varies by zone",
    x        = NULL,
    y        = "Mean wind speed (m/s)",
    fill     = NULL,
    caption  = "Source: Copernicus ERA5 reanalysis"
  )
```


### 3.3 Solar irradiance gradient {#sec-cross-zone-solar}

Surface solar irradiance falls with latitude — southern zones receive
materially more annual radiation than Nordic zones. This gradient drives
differences in solar PV economics and in the relative contribution of
solar variability to price risk across the interconnected market.


```{r}
library(plotly)

solar_diurnal <- era5_all |>
  mutate(
    hour_utc = hour(datetime_utc),
    season = case_when(
      month(datetime_utc) %in% c(12, 1, 2) ~ "Winter",
      month(datetime_utc) %in% c(3, 4, 5)  ~ "Spring",
      month(datetime_utc) %in% c(6, 7, 8)  ~ "Summer",
      TRUE                                  ~ "Autumn"
    ),
    season = factor(season, levels = c("Winter", "Spring", "Summer", "Autumn"))
  ) |>
  group_by(zone, season, hour_utc) |>
  summarise(mean_ssrd = round(mean(ssrd_wm2, na.rm = TRUE), 1),
            .groups = "drop")

make_panel <- function(df, season_label, show_legend = FALSE) {
  plot_ly(df |> filter(season == season_label),
          x = ~hour_utc, y = ~mean_ssrd, color = ~zone,
          type = "scatter", mode = "lines",
          showlegend = show_legend,
          hovertemplate = paste0(
            "<b>%{meta}</b><br>",
            "Hour: %{x} UTC<br>",
            "Irradiance: %{y:.1f} W/m²<extra></extra>"
          ),
          meta = ~zone) |>
    layout(
      xaxis = list(title = if (season_label %in% c("Summer", "Autumn"))
                             "Hour (UTC)" else ""),
      yaxis = list(title = if (season_label %in% c("Winter", "Summer"))
                             "W/m²" else ""),
      annotations = list(
        list(text = paste0("<b>", season_label, "</b>"),
             xref = "paper", yref = "paper",
             x = 0.5, y = 1.05, showarrow = FALSE,
             font = list(size = 12))
      )
    )
}

subplot(
  make_panel(solar_diurnal, "Winter"),
  make_panel(solar_diurnal, "Spring", show_legend = TRUE),
  make_panel(solar_diurnal, "Summer"),
  make_panel(solar_diurnal, "Autumn"),
  nrows = 2, shareX = TRUE, shareY = TRUE,
  titleX = TRUE, titleY = TRUE
) |>
  layout(
    title = list(
      text = "Mean diurnal solar irradiance by zone and season",
      font = list(size = 14)
    ),
    legend = list(title = list(text = "Zone")),
    annotations = list(
      list(text = "Source: Copernicus ERA5 reanalysis",
           xref = "paper", yref = "paper", x = 1, y = -0.08,
           showarrow = FALSE, font = list(size = 8, color = "grey50"),
           xanchor = "right")
    )
  )
```



### 3.4 Cross-zone summary statistics {#sec-cross-zone-summary}

```{r}
#| label: tbl-cross-zone-summary

era5_all |>
  group_by(zone) |>
  summarise(
    `Wind 100m mean (m/s)` = round(mean(wind_speed_100m, na.rm = TRUE), 2),
    `Wind 100m sd (m/s)`   = round(sd(wind_speed_100m, na.rm = TRUE), 2),
    `Solar mean (W/m²)`    = round(mean(ssrd_wm2, na.rm = TRUE), 1),
    `Solar daylight mean (W/m²)` = round(
      mean(ssrd_wm2[ssrd_wm2 > 0], na.rm = TRUE), 1
    ),
    `Temp mean (°C)`       = round(mean(temperature_2m, na.rm = TRUE), 1),
    `Hours`                = n(),
    .groups = "drop"
  ) |>
  knitr::kable(caption = "ERA5 weather summary statistics by bidding zone")
```

---

*The remainder of this page focuses on DE-LU, where SMARD provides
hourly generation by technology (wind onshore, wind offshore, solar PV)
and BNetzA publishes installed-capacity milestones — the data needed to
compute capacity factors and fit weather → generation models.*


## 4.0 DE-LU weather profiles {#sec-delu-eda}

### 4.1 Distributions {#sec-delu-distributions}

```{r}
#| label: fig-delu-distributions
#| fig-height: 7
#| fig-width: 10

p_wind100 <- ggplot(era5, aes(x = wind_speed_100m)) +
  geom_histogram(bins = 80, fill = pal["wind_onshore"], alpha = 1.7) +
  labs(title = "100 m wind speed", x = "m/s", y = "Count")

p_wind10 <- ggplot(era5, aes(x = wind_speed_10m)) +
  geom_histogram(bins = 80, fill = pal["wind_offshore"], alpha = 1.7) +
  labs(title = "10 m wind speed", x = "m/s", y = "Count")

p_temp <- ggplot(era5, aes(x = temperature_2m)) +
  geom_histogram(bins = 80, fill = "grey50", alpha = 1.7) +
  labs(title = "2 m temperature", x = "°C", y = "Count")

p_ssrd <- era5 |>
  filter(ssrd_wm2 > 0) |>
  ggplot(aes(x = ssrd_wm2)) +
  geom_histogram(bins = 80, fill = pal["solar"], alpha = 1.7) +
  labs(title = "Surface solar radiation (daylight only)",
       x = expression("W/m"^2), y = "Count")

(p_wind100 + p_wind10) / (p_temp + p_ssrd) +
  plot_annotation(
    title   = glue("ERA5 weather variable distributions — {zone_file}"),
    caption = "Source: Copernicus ERA5 reanalysis"
  )
```


### 4.2 Seasonal diurnal profiles {#sec-delu-diurnal}

```{r}
#| label: fig-delu-diurnal
#| fig-height: 6
#| fig-width: 10

era5 |>
  mutate(
    hour_utc = hour(datetime_utc),
    season = case_when(
      month(datetime_utc) %in% c(12, 1, 2) ~ "Winter",
      month(datetime_utc) %in% c(3, 4, 5)  ~ "Spring",
      month(datetime_utc) %in% c(6, 7, 8)  ~ "Summer",
      TRUE                                  ~ "Autumn"
    ),
    season = factor(season, levels = c("Winter", "Spring", "Summer", "Autumn"))
  ) |>
  group_by(season, hour_utc) |>
  summarise(
    wind_100m = mean(wind_speed_100m, na.rm = TRUE),
    ssrd      = mean(ssrd_wm2, na.rm = TRUE),
    .groups   = "drop"
  ) |>
  pivot_longer(c(wind_100m, ssrd), names_to = "variable", values_to = "value") |>
  mutate(variable = recode(variable,
                           wind_100m = "Wind speed 100m (m/s)",
                           ssrd      = "Solar radiation (W/m²)")) |>
  ggplot(aes(x = hour_utc, y = value, colour = season)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 1.2) +
  facet_wrap(~ variable, scales = "free_y") +
  scale_colour_brewer(palette = "Set2") +
  labs(
    title   = glue("Mean diurnal profiles by season — {zone_file}"),
    x       = "Hour (UTC)",
    y       = NULL,
    colour  = NULL,
    caption = "Source: Copernicus ERA5 reanalysis"
  )
```


### 4.3 Timestamp alignment check {#sec-timestamp-check}

ERA5 `ssrd` is originally an accumulated field (J/m² per forecast step).
Before modelling we verify that the converted instantaneous values align
temporally with SMARD solar generation — a 1-hour offset would produce
misleading scatter plots and biased models.

```{r}
#| label: fig-timestamp-alignment
#| fig-height: 4

era5 |>
  inner_join(smard, by = "datetime_utc") |>
  filter(month(datetime_utc) == 6) |>
  mutate(hour_utc = hour(datetime_utc)) |>
  filter(hour_utc %in% 4:20) |>
  group_by(hour_utc) |>
  summarise(
    mean_ssrd  = mean(ssrd_wm2, na.rm = TRUE),
    mean_solar = mean(solar, na.rm = TRUE),
    .groups    = "drop"
  ) |>
  pivot_longer(-hour_utc, names_to = "source", values_to = "value") |>
  mutate(source = recode(source,
                         mean_ssrd  = "ERA5 solar radiation",
                         mean_solar = "SMARD solar generation")) |>
  ggplot(aes(x = hour_utc, y = value, colour = source)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  facet_wrap(~ source, scales = "free_y") +
  labs(
    title    = "Timestamp alignment: ERA5 vs SMARD (June)",
    subtitle = "Peaks at the same UTC hour confirm correct alignment",
    x        = "Hour (UTC)",
    y        = NULL,
    colour   = NULL,
    caption  = "Sources: Copernicus ERA5, SMARD (BNetzA)"
  ) +
  theme(legend.position = "none")
```


## 5.0 Join & capacity factors {#sec-capacity}

```{r}
#| label: join-datasets

model_raw <- era5 |>
  inner_join(smard, by = "datetime_utc") |>
  mutate(
    date  = as_date(datetime_utc),
    month = factor(month(datetime_utc, label = TRUE), ordered = FALSE),
    hour  = factor(hour(datetime_utc)),
    year  = year(datetime_utc),
    season = case_when(
      month(datetime_utc) %in% c(12, 1, 2) ~ "Winter",
      month(datetime_utc) %in% c(3, 4, 5)  ~ "Spring",
      month(datetime_utc) %in% c(6, 7, 8)  ~ "Summer",
      TRUE                                  ~ "Autumn"
    ),
    season = factor(season, levels = c("Winter", "Spring", "Summer", "Autumn"))
  )

cat("Joined rows:", nrow(model_raw), "\n")
```

Computing capacity factors requires dividing observed generation (MW) by
installed capacity (MW). SMARD does not include an installed-capacity
column, so we construct a lookup from
[BNetzA / Marktstammdatenregister](https://www.marktstammdatenregister.de)
year-end published totals and interpolate linearly to daily resolution.

```{r}
#| label: installed-capacity

# ── Year-end installed capacity (MW) — BNetzA monitoring reports ─────
# Sources: BNetzA Monitoringbericht, SMARD "Installed generation capacity"
# These are approximate — replace with exact values if available.

installed_cap <- tribble(
  ~date,            ~cap_wind_onshore, ~cap_wind_offshore, ~cap_solar,
  "2019-12-31",     53200,             7700,               49100,
  "2020-12-31",     54400,             7770,               53700,
  "2021-12-31",     55900,             7790,               58700,
  "2022-12-31",     58100,             8050,               66500,
  "2023-12-31",     61000,             8390,               74100,
  "2024-12-31",     63800,             8900,               81700,
  "2025-06-30",     65500,             9200,               86000   # provisional
) |>
  mutate(date = as_date(date))

# ── Linearly interpolate to daily resolution ─────────────────────────
date_seq <- tibble(date = seq(min(installed_cap$date),
                              max(installed_cap$date),
                              by = "day"))

cap_daily <- date_seq |>
  left_join(installed_cap, by = "date") |>
  mutate(across(starts_with("cap_"),
                ~ approx(x = date[!is.na(.x)],
                          y = .x[!is.na(.x)],
                          xout = date)$y))
```

```{r}
#| label: compute-cf

model_data <- model_raw |>
  left_join(cap_daily, by = "date") |>
  mutate(
    cf_wind_onshore  = wind_onshore  / cap_wind_onshore,
    cf_wind_offshore = wind_offshore / cap_wind_offshore,
    cf_solar         = solar         / cap_solar
  ) |>
  # Solar CF is only meaningful during daylight; set nighttime to NA
  mutate(cf_solar = if_else(ssrd_wm2 > 0 | solar > 0,
                            cf_solar,
                            NA_real_))

cat("Observations with valid CF:",
    sum(!is.na(model_data$cf_wind_onshore)), "(wind on),",
    sum(!is.na(model_data$cf_wind_offshore)), "(wind off),",
    sum(!is.na(model_data$cf_solar)), "(solar)\n")
```


## 6.0 Correlation structure {#sec-correlations}

```{r}
#| label: fig-correlation
#| fig-height: 6

cor_vars <- model_data |>
  select(
    `Wind 100m`   = wind_speed_100m,
    `Wind 10m`    = wind_speed_10m,
    `Temp 2m`     = temperature_2m,
    `Solar rad`   = ssrd_wm2,
    `CF onshore`  = cf_wind_onshore,
    `CF offshore` = cf_wind_offshore,
    `CF solar`    = cf_solar
  ) |>
  drop_na()

cor_mat <- cor(cor_vars)

cor_long <- cor_mat |>
  as.data.frame() |>
  rownames_to_column("var1") |>
  pivot_longer(-var1, names_to = "var2", values_to = "r") |>
  mutate(
    var1 = factor(var1, levels = colnames(cor_mat)),
    var2 = factor(var2, levels = colnames(cor_mat))
  ) |>
  filter(as.integer(var1) >= as.integer(var2))

ggplot(cor_long, aes(x = var2, y = var1, fill = r)) +
  geom_tile(colour = "white", linewidth = 0.5) +
  geom_text(aes(label = sprintf("%.2f", r)), size = 3.2) +
  scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B",
                       midpoint = 0, limits = c(-1, 1)) +
  labs(
    title    = "Weather–generation correlation matrix",
    subtitle = glue("{zone_file}, hourly observations"),
    fill     = "Pearson r",
    caption  = "Sources: Copernicus ERA5, SMARD (BNetzA)"
  ) +
  coord_equal() +
  theme(axis.title = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1))
```


## 7.0 Scatter plots: weather → capacity factor {#sec-scatter}

### 7.1 Wind speed → wind onshore CF {#sec-wind-scatter}

```{r}
#| label: fig-wind-scatter
#| fig-height: 5

model_data |>
  filter(!is.na(cf_wind_onshore)) |>
  slice_sample(n = 15000) |>
  ggplot(aes(x = wind_speed_100m, y = cf_wind_onshore)) +
  geom_point(alpha = 0.15, colour = pal["wind_onshore"], size = 0.5) +
  geom_smooth(method = "lm",
              formula = y ~ poly(x, 3),
              colour = "black", linewidth = 0.8, se = TRUE) +
  scale_y_continuous(labels = scales::percent_format(),
                     limits = c(0, 1)) +
  labs(
    title    = "Wind speed → onshore capacity factor",
    subtitle = glue("Degree-3 polynomial fit | {zone_file} hourly"),
    x        = "ERA5 100 m wind speed (m/s)",
    y        = "Capacity factor",
    caption  = "Sources: Copernicus ERA5, SMARD (BNetzA)"
  )
```


### 7.2 Solar irradiance → solar PV CF {#sec-solar-scatter}

The marginal relationship between irradiance and solar CF appears
non-monotonic when plotted unconditionally. This is not a modelling
problem — it reflects the mixing of seasonal regimes. At the same
irradiance level (e.g. 100 W/m²), winter midday produces a high CF
(that is as good as the sun gets in January) while summer cloud cover
at the same irradiance implies suppressed output relative to clear-sky
potential. The hour and month fixed effects in the model capture this
conditioning; the faceted plots below confirm monotonic relationships
within time-of-day strata.

```{r}
#| label: fig-solar-scatter-marginal
#| fig-height: 8
#| fig-width: 12

solar_plot_df <- model_data |>
  filter(!is.na(cf_solar), ssrd_wm2 > 5)

p_marginal <- solar_plot_df |>
  slice_sample(n = 15000) |>
  ggplot(aes(x = ssrd_wm2, y = cf_solar)) +
  geom_point(alpha = 0.15, colour = pal["solar"], size = 0.5) +
  geom_smooth(method = "gam", formula = y ~ s(x),
              colour = "black", linewidth = 0.8, se = TRUE) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Marginal (unconditional)",
    x     = expression("W/m"^2),
    y     = "Capacity factor"
  )

p_faceted <- solar_plot_df |>
  mutate(hour_bin = cut(hour(datetime_utc),
                        breaks = seq(4, 22, by = 2),
                        labels = paste0(seq(4, 20, by = 2), "–",
                                        seq(6, 22, by = 2)),
                        include.lowest = TRUE)) |>
  filter(!is.na(hour_bin)) |>
  slice_sample(n = 20000) |>
  ggplot(aes(x = ssrd_wm2, y = cf_solar)) +
  geom_point(alpha = 0.16, colour = pal["solar"], size = 0.3) +
  geom_smooth(method = "gam", formula = y ~ s(x),
              colour = "black", linewidth = 0.7) +
  facet_wrap(~ hour_bin, ncol = 3) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Conditioned on time of day (UTC)",
    x     = expression("W/m"^2),
    y     = "Capacity factor"
  )

p_marginal | p_faceted +
  plot_annotation(
    title   = "Solar irradiance → PV capacity factor",
    caption = "Sources: Copernicus ERA5, SMARD (BNetzA)"
  )
```


### 7.3 Midday deep dive: season resolves the remaining non-monotonicity {#sec-solar-midday}

Even within the 10–12 UTC bin, the relationship appears non-monotonic.
Faceting by season confirms this is a **solar geometry effect**: the
same irradiance value implies very different things in winter versus
summer.

```{r}
#| label: fig-solar-midday-season
#| fig-height: 5
#| fig-width: 10

model_data |>
  filter(!is.na(cf_solar), ssrd_wm2 > 5,
         hour(datetime_utc) %in% 10:11) |>
  ggplot(aes(x = ssrd_wm2, y = cf_solar)) +
  geom_point(alpha = 0.18, colour = pal["solar"], size = 0.5) +
  geom_smooth(method = "gam", formula = y ~ s(x),
              colour = "black", linewidth = 0.8) +
  facet_wrap(~ season) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title    = "10–12 UTC: irradiance → CF by season",
    subtitle = "Within-season monotonicity confirms need for month × irradiance interactions",
    x        = expression("W/m"^2),
    y        = "Capacity factor",
    caption  = "Sources: Copernicus ERA5, SMARD (BNetzA)"
  )
```


## 8.0 Capacity factor models {#sec-models}

We fit one `tidymodels` workflow per technology using a temporal
80/20 train/test split (no shuffling — critical for time-series
validity). Wind models use polynomial wind-speed terms with month/hour
fixed effects. The solar model includes **month × irradiance
interaction terms** to capture the seasonal variation in the
irradiance–CF slope identified in the EDA above.


### 8.1 Wind onshore {#sec-model-wind-on}

```{r}
#| label: model-wind-onshore

wind_on_df <- model_data |>
  filter(!is.na(cf_wind_onshore)) |>
  select(datetime_utc, cf_wind_onshore,
         wind_speed_100m, wind_speed_10m, temperature_2m,
         month, hour)

split_wind_on <- initial_time_split(wind_on_df, prop = 0.8)
cat("Train:", nrow(training(split_wind_on)),
    "| Test:", nrow(testing(split_wind_on)), "\n")

rec_wind_on <- recipe(
  cf_wind_onshore ~ wind_speed_100m + wind_speed_10m +
                     temperature_2m + month + hour,
  data = training(split_wind_on)
) |>
  step_poly(wind_speed_100m, degree = 3) |>
  step_poly(wind_speed_10m,  degree = 2) |>
  step_dummy(all_nominal_predictors())

wf_wind_on <- workflow() |>
  add_recipe(rec_wind_on) |>
  add_model(linear_reg())

fit_wind_on <- fit(wf_wind_on, data = training(split_wind_on))

preds_wind_on <- fit_wind_on |>
  augment(testing(split_wind_on))

metrics_wind_on <- preds_wind_on |>
  metrics(truth = cf_wind_onshore, estimate = .pred)

metrics_wind_on |>
  knitr::kable(digits = 4, caption = "Wind onshore model — test set performance")
```


### 8.2 Wind offshore {#sec-model-wind-off}

```{r}
#| label: model-wind-offshore

wind_off_df <- model_data |>
  filter(!is.na(cf_wind_offshore)) |>
  select(datetime_utc, cf_wind_offshore,
         wind_speed_100m, wind_speed_10m, temperature_2m,
         month, hour)

split_wind_off <- initial_time_split(wind_off_df, prop = 0.8)

rec_wind_off <- recipe(
  cf_wind_offshore ~ wind_speed_100m + wind_speed_10m +
                      temperature_2m + month + hour,
  data = training(split_wind_off)
) |>
  step_poly(wind_speed_100m, degree = 3) |>
  step_poly(wind_speed_10m,  degree = 2) |>
  step_dummy(all_nominal_predictors())

wf_wind_off <- workflow() |>
  add_recipe(rec_wind_off) |>
  add_model(linear_reg())

fit_wind_off <- fit(wf_wind_off, data = training(split_wind_off))

preds_wind_off <- fit_wind_off |>
  augment(testing(split_wind_off))

metrics_wind_off <- preds_wind_off |>
  metrics(truth = cf_wind_offshore, estimate = .pred)

metrics_wind_off |>
  knitr::kable(digits = 4, caption = "Wind offshore model — test set performance")
```

::: {.callout-warning title="Offshore wind model — known limitation"}
The offshore wind model achieves an R² of only ~0.29 on the test set —
substantially weaker than onshore wind (R² ~0.92) or solar PV (R² ~0.84).
This is a spatial resolution problem, not a specification issue: ERA5 wind
speeds are averaged over the entire DE-LU zone using cosine-latitude
weighting, which dilutes the coastal and maritime wind signals that drive
offshore turbine output. Offshore farms in the German North Sea and Baltic
experience wind regimes that differ meaningfully from the zone-average
conditions captured by the ERA5 spatial mean.

**Downstream implications.** This model is used in Stage 3 to predict
offshore capacity factors for the Monte Carlo revenue simulation. At
R² = 0.29, the predictions are essentially noisy — they will understate
the true range of offshore generation outcomes. We flag this as a known
limitation of the current data pipeline. A targeted improvement would be
to extract ERA5 grid cells overlapping German offshore wind farm clusters
(e.g. the Helgoland/Borkum cluster in the North Sea) rather than using
the full DE-LU spatial average. For this iteration, the Stage 3 portfolio
analysis focuses on onshore wind where the CF model is reliable.
:::


### 8.3 Solar PV {#sec-model-solar}

The solar recipe includes month × irradiance interactions so that the
irradiance–CF slope can vary by month — physically motivated by the
seasonal change in solar geometry (sun angle, day length) identified in
@sec-solar-midday. Temperature is included to capture the ~0.4%/°C
efficiency loss above standard test conditions (25 °C).

```{r}
#| label: model-solar

solar_df <- model_data |>
  filter(!is.na(cf_solar), ssrd_wm2 > 0) |>
  select(datetime_utc, cf_solar,
         ssrd_wm2, temperature_2m,
         month, hour)

split_solar <- initial_time_split(solar_df, prop = 0.8)

rec_solar <- recipe(
  cf_solar ~ ssrd_wm2 + temperature_2m + month + hour,
  data = training(split_solar)
) |>
  step_dummy(all_nominal_predictors()) |>
  step_interact(terms = ~ ssrd_wm2:starts_with("month_"))

wf_solar <- workflow() |>
  add_recipe(rec_solar) |>
  add_model(linear_reg())

fit_solar <- fit(wf_solar, data = training(split_solar))

preds_solar <- fit_solar |>
  augment(testing(split_solar))

metrics_solar <- preds_solar |>
  metrics(truth = cf_solar, estimate = .pred)

metrics_solar |>
  knitr::kable(digits = 4, caption = "Solar PV model — test set performance")
```


## 9.0 Model diagnostics {#sec-diagnostics}

Two diagnostic panels per technology: predicted versus actual capacity
factor (a well-calibrated model hugs the 1:1 line) and the residual
distribution (centred near zero with no heavy skew indicates the model
specification is adequate).

```{r}
#| label: fig-diagnostics
#| fig-height: 10
#| fig-width: 10

diag_panels <- function(preds, truth_col, label, colour) {
  n_plot <- min(8000, nrow(preds))

  p_scatter <- preds |>
    slice_sample(n = n_plot) |>
    ggplot(aes(x = .data[[truth_col]], y = .pred)) +
    geom_point(alpha = 0.15, colour = colour, size = 0.5) +
    geom_abline(slope = 1, intercept = 0,
                linetype = "dashed", colour = "grey30") +
    scale_x_continuous(labels = scales::percent_format()) +
    scale_y_continuous(labels = scales::percent_format()) +
    coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
    labs(title = label, x = "Actual CF", y = "Predicted CF")

  p_resid <- preds |>
    mutate(resid = .pred - .data[[truth_col]]) |>
    ggplot(aes(x = resid)) +
    geom_histogram(bins = 80, fill = colour, alpha = 0.7) +
    geom_vline(xintercept = 0, linetype = "dashed", colour = "grey30") +
    labs(title = paste(label, "— residuals"),
         x = "Residual (pred − actual)", y = "Count")

  p_scatter + p_resid
}

p1 <- diag_panels(preds_wind_on,  "cf_wind_onshore",  "Wind onshore",  pal["wind_onshore"])
p2 <- diag_panels(preds_wind_off, "cf_wind_offshore", "Wind offshore", pal["wind_offshore"])
p3 <- diag_panels(preds_solar,    "cf_solar",         "Solar PV",      pal["solar"])

(p1 / p2 / p3) +
  plot_annotation(
    title   = glue("Capacity factor model diagnostics — {zone_file} test set"),
    caption = "Sources: Copernicus ERA5, SMARD (BNetzA)"
  )
```


## 9.1 Residual autocorrelation {#sec-acf}

The scatter and histogram diagnostics above assess bias and distributional
shape, but not temporal dependence. If model residuals are strongly
autocorrelated, it indicates systematic patterns the specification is
missing — for instance, multi-day weather persistence or seasonal ramp
events that the polynomial + fixed effects structure cannot capture.

We plot the ACF and PACF for the wind onshore and solar PV test-set
residuals (offshore is omitted given its low R²). Significant
autocorrelation at short lags (1–6 hours) is expected — weather
persistence is physical — but decay beyond ~24 hours would suggest the
model is adequate for the hourly predictions used in Stage 3.

```{r}
#| label: fig-acf-pacf
#| fig-cap: "Residual ACF and PACF for the wind onshore and solar PV capacity factor models (test set). Significant short-lag autocorrelation reflects physical weather persistence; rapid decay beyond 24–48 hours indicates the model captures the dominant temporal structure."
#| fig-width: 10
#| fig-height: 8

acf_plot <- function(preds, truth_col, label, colour, max_lag = 72) {
  resids <- preds |>
    arrange(datetime_utc) |>
    mutate(resid = .pred - .data[[truth_col]]) |>
    pull(resid)

  acf_vals  <- acf(resids, lag.max = max_lag, plot = FALSE)
  pacf_vals <- pacf(resids, lag.max = max_lag, plot = FALSE)

  ci <- qnorm(0.975) / sqrt(length(resids))

  p_acf <- tibble(lag = acf_vals$lag[-1], acf = acf_vals$acf[-1]) |>
    ggplot(aes(x = lag, y = acf)) +
    geom_hline(yintercept = c(-ci, ci), linetype = "dashed", colour = "grey50") +
    geom_hline(yintercept = 0, colour = "grey30") +
    geom_segment(aes(xend = lag, yend = 0), colour = colour, linewidth = 0.5) +
    labs(title = paste(label, "— ACF"), x = "Lag (hours)", y = "ACF")

  p_pacf <- tibble(lag = pacf_vals$lag, pacf = pacf_vals$acf) |>
    ggplot(aes(x = lag, y = pacf)) +
    geom_hline(yintercept = c(-ci, ci), linetype = "dashed", colour = "grey50") +
    geom_hline(yintercept = 0, colour = "grey30") +
    geom_segment(aes(xend = lag, yend = 0), colour = colour, linewidth = 0.5) +
    labs(title = paste(label, "— PACF"), x = "Lag (hours)", y = "PACF")

  p_acf + p_pacf
}

p_acf_wind <- acf_plot(preds_wind_on, "cf_wind_onshore", "Wind onshore", pal["wind_onshore"])
p_acf_solar <- acf_plot(preds_solar, "cf_solar", "Solar PV", pal["solar"])

(p_acf_wind / p_acf_solar) +
  plot_annotation(
    title   = glue("Residual autocorrelation — {zone_file} CF models (test set)"),
    subtitle = "Dashed lines = 95% confidence bounds under white noise null",
    caption = "Sources: Copernicus ERA5, SMARD (BNetzA)"
  )
```

::: {.callout-note title="Interpreting the ACF/PACF"}
Significant autocorrelation at lags 1–6 is expected and not concerning for
this application: weather systems persist for hours, so a model that
conditions on instantaneous weather will inevitably leave correlated
residuals. What matters for downstream use is whether the residuals exhibit
*systematic* multi-day structure. If the ACF decays to near zero by lag 48,
the model is capturing the dominant temporal features and the remaining
autocorrelation reflects irreducible weather persistence. If the PACF shows
spikes at 24-hour intervals, a diurnal component may be under-specified —
but this is unlikely given the hour fixed effects already in the recipe.
:::


## 10.0 Save fitted workflows {#sec-export}

The fitted `tidymodels` workflow objects are serialised to `models/` for
reuse in Stage 2 (price decomposition) and Stage 3 (risk simulation).
Each `.rds` file contains the full preprocessing recipe plus fitted
coefficients — `predict(readRDS(...), new_data)` is all downstream
stages need.

```{r}
#| label: save-models

dir.create("models", showWarnings = FALSE)

saveRDS(fit_wind_on,  glue("models/cf_wind_onshore_{zone_file}.rds"))
saveRDS(fit_wind_off, glue("models/cf_wind_offshore_{zone_file}.rds"))
saveRDS(fit_solar,    glue("models/cf_solar_{zone_file}.rds"))

cat(glue("Saved 3 workflow objects to models/ ({zone_file})"), "\n")
```


## 11.0 Key findings {#sec-findings}

```{r}
#| label: summary-table

bind_rows(
  metrics_wind_on  |> mutate(technology = "Wind onshore"),
  metrics_wind_off |> mutate(technology = "Wind offshore"),
  metrics_solar    |> mutate(technology = "Solar PV")
) |>
  select(technology, .metric, .estimate) |>
  pivot_wider(names_from = .metric, values_from = .estimate) |>
  knitr::kable(
    digits = 4,
    caption = glue("Summary of capacity factor model performance on the temporal test set ({zone_file})")
  )
```

**Cross-zone context.** The ERA5 weather profiles in @sec-cross-zone
reveal substantial geographic variation in renewable resource quality.
Wind speeds are highest in coastal and Nordic zones, while solar
irradiance follows a clear latitude gradient. These physical differences
drive the cross-zone variation in renewable generation patterns — and
ultimately in the price dynamics — explored in Stage 2.

**Derived weather features.** HDD and CDD (@sec-hdd-cdd) translate ERA5
temperature into demand-relevant signals for Stage 2. Germany's climate
is heating-dominated: HDD values are substantially larger and more
variable than CDD, consistent with winter load being the primary
temperature-driven demand component in the DE-LU zone.

**DE-LU capacity factor models.** ERA5 100 m wind speed is the dominant
predictor of wind generation capacity factors, with the cubic polynomial
capturing the characteristic ramp-up and saturation of a turbine power
curve. Solar irradiance maps onto PV capacity factors with a slope that
varies by month — an interaction effect driven by solar geometry (sun
angle and day length). The month × irradiance terms resolve the apparent
non-monotonicity visible in the unconditional scatter plot
(@sec-solar-scatter), confirming that the bimodal pattern reflects
seasonal regime mixing rather than a data quality issue. Temperature
adds a modest correction for PV cell-efficiency losses above standard
test conditions (25 °C).

**Offshore wind limitation.** The offshore CF model (R² ~0.29) is
substantially weaker than onshore or solar, due to the spatial averaging
of ERA5 over the full DE-LU zone diluting coastal wind signals. This is
a known limitation flagged in @sec-model-wind-off; the Stage 3 portfolio
analysis focuses on onshore wind where the CF model is reliable.

**Residual autocorrelation.** The ACF/PACF analysis (@sec-acf) confirms
that short-lag autocorrelation in CF model residuals reflects physical
weather persistence rather than model misspecification. Rapid decay
beyond 24–48 hours indicates the polynomial + fixed effects specification
captures the dominant temporal structure.

These fitted workflows feed directly into Stage 2, where weather-driven
generation predictions are used as features in the day-ahead price
decomposition model.
