---
title: "Weather → Generation"
subtitle: "Capacity factor models linking ERA5 reanalysis to observed renewable output"
params:
  bidding_zone: "NL"
  analysis_year_start: 2020
  analysis_year_end: 2025
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 9,
  fig.height = 5
)

library(tidyverse)
library(tidymodels)
library(tsibble)
library(arrow)
library(patchwork)
library(lubridate)
library(glue)

# ── Zone label for filenames and display ─────────────────────────────
zone_file <- str_replace(params$bidding_zone, "_", "-")  # DE_LU → DE-LU

# ── Custom theme (consistent with 01_data.qmd) ──────────────────────
theme_report <- theme_minimal(base_family = "Arial", base_size = 12) +
  theme(
    plot.title    = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(colour = "grey40", size = 11),
    plot.caption  = element_text(colour = "grey50", size = 8),
    strip.text    = element_text(face = "bold"),
    legend.position = "bottom"
  )

theme_set(theme_report)

# ── Colour palette ───────────────────────────────────────────────────
pal <- c(
  wind_onshore  = "#2166AC",
  wind_offshore = "#67A9CF",
  solar         = "#F4A582"
)
```


## 1.0 Overview {#sec-overview}

This page establishes the physical link between weather conditions and
renewable electricity generation — the foundation every subsequent stage
depends on. We fit capacity-factor models for **wind onshore**, **wind
offshore**, and **solar PV** in the `r zone_file` bidding zone using ERA5
reanalysis weather data and observed SMARD generation series.

**Approach.** For each technology we:

1. Profile the ERA5 weather variables and check timestamp alignment with
   SMARD generation.
2. Join hourly ERA5 weather to observed generation and compute capacity
   factors using BNetzA installed-capacity milestones.
3. Fit a `tidymodels` workflow — polynomial weather terms plus
   month/hour fixed effects (with month × irradiance interactions for
   solar) — on a temporal 80/20 train/test split.
4. Evaluate with RMSE, MAE, and R² on the held-out test period.


## 2.0 Data loading {#sec-data}

```{r}
#| label: load-data

era5_path  <- glue("data/processed/era5/{zone_file}.parquet")
smard_path <- list.files("data/processed/smard",
                         pattern = glue("smard_{zone_file}"),
                         full.names = TRUE)

era5  <- read_parquet(era5_path)
smard <- read_parquet(smard_path)

cat("ERA5  rows:", nrow(era5),
    " | range:", format(min(era5$datetime_utc)), "–",
    format(max(era5$datetime_utc)), "\n")
cat("SMARD rows:", nrow(smard),
    " | range:", format(min(smard$datetime_utc)), "–",
    format(max(smard$datetime_utc)), "\n")
```


## 3.0 ERA5 weather variable profiles {#sec-era5-eda}

Before joining to generation data, we profile the four ERA5 variables to
check for processing artifacts, confirm physical ranges, and understand
the marginal distributions.

### 3.1 Distributions {#sec-era5-distributions}

```{r}
#| label: fig-era5-distributions
#| fig-height: 7
#| fig-width: 10

p_wind100 <- ggplot(era5, aes(x = wind_speed_100m)) +
  geom_histogram(bins = 80, fill = pal["wind_onshore"], alpha = 0.7) +
  labs(title = "100 m wind speed", x = "m/s", y = "Count")

p_wind10 <- ggplot(era5, aes(x = wind_speed_10m)) +
  geom_histogram(bins = 80, fill = pal["wind_offshore"], alpha = 0.7) +
  labs(title = "10 m wind speed", x = "m/s", y = "Count")

p_temp <- ggplot(era5, aes(x = temperature_2m)) +
  geom_histogram(bins = 80, fill = "grey50", alpha = 0.7) +
  labs(title = "2 m temperature", x = "°C", y = "Count")

p_ssrd <- era5 |>
  filter(ssrd_wm2 > 0) |>
  ggplot(aes(x = ssrd_wm2)) +
  geom_histogram(bins = 80, fill = pal["solar"], alpha = 0.7) +
  labs(title = "Surface solar radiation (daylight only)",
       x = expression("W/m"^2), y = "Count")

(p_wind100 + p_wind10) / (p_temp + p_ssrd) +
  plot_annotation(
    title   = glue("ERA5 weather variable distributions — {zone_file} spatial average"),
    caption = "Source: Copernicus ERA5 reanalysis"
  )
```

### 3.2 Seasonal diurnal profiles {#sec-era5-diurnal}

```{r}
#| label: fig-era5-diurnal
#| fig-height: 6
#| fig-width: 10

era5 |>
  mutate(
    hour_utc = hour(datetime_utc),
    season = case_when(
      month(datetime_utc) %in% c(12, 1, 2) ~ "Winter",
      month(datetime_utc) %in% c(3, 4, 5)  ~ "Spring",
      month(datetime_utc) %in% c(6, 7, 8)  ~ "Summer",
      TRUE                                  ~ "Autumn"
    ),
    season = factor(season, levels = c("Winter", "Spring", "Summer", "Autumn"))
  ) |>
  group_by(season, hour_utc) |>
  summarise(
    wind_100m = mean(wind_speed_100m, na.rm = TRUE),
    ssrd      = mean(ssrd_wm2, na.rm = TRUE),
    .groups   = "drop"
  ) |>
  pivot_longer(c(wind_100m, ssrd), names_to = "variable", values_to = "value") |>
  mutate(variable = recode(variable,
                           wind_100m = "Wind speed 100m (m/s)",
                           ssrd      = "Solar radiation (W/m²)")) |>
  ggplot(aes(x = hour_utc, y = value, colour = season)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 1.2) +
  facet_wrap(~ variable, scales = "free_y") +
  scale_colour_brewer(palette = "Set2") +
  labs(
    title   = "Mean diurnal profiles by season",
    x       = "Hour (UTC)",
    y       = NULL,
    colour  = NULL,
    caption = "Source: Copernicus ERA5 reanalysis"
  )
```


### 3.3 Timestamp alignment check {#sec-timestamp-check}

ERA5 `ssrd` is originally an accumulated field (J/m² per forecast step).
Before modelling we verify that the converted instantaneous values align
temporally with SMARD solar generation — a 1-hour offset would produce
misleading scatter plots and biased models.

```{r}
#| label: fig-timestamp-alignment
#| fig-height: 4

era5 |>
  inner_join(smard, by = "datetime_utc") |>
  filter(month(datetime_utc) == 6) |>
  mutate(hour_utc = hour(datetime_utc)) |>
  filter(hour_utc %in% 4:20) |>
  group_by(hour_utc) |>
  summarise(
    mean_ssrd  = mean(ssrd_wm2, na.rm = TRUE),
    mean_solar = mean(solar, na.rm = TRUE),
    .groups    = "drop"
  ) |>
  pivot_longer(-hour_utc, names_to = "source", values_to = "value") |>
  mutate(source = recode(source,
                         mean_ssrd  = "ERA5 solar radiation",
                         mean_solar = "SMARD solar generation")) |>
  ggplot(aes(x = hour_utc, y = value, colour = source)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  facet_wrap(~ source, scales = "free_y") +
  labs(
    title    = "Timestamp alignment: ERA5 vs SMARD (June)",
    subtitle = "Peaks at the same UTC hour confirm correct alignment",
    x        = "Hour (UTC)",
    y        = NULL,
    colour   = NULL,
    caption  = "Sources: Copernicus ERA5, SMARD (BNetzA)"
  ) +
  theme(legend.position = "none")
```


## 4.0 Join & capacity factors {#sec-capacity}

```{r}
#| label: join-datasets

model_raw <- era5 |>
  inner_join(smard, by = "datetime_utc") |>
  mutate(
    date  = as_date(datetime_utc),
    month = factor(month(datetime_utc, label = TRUE), ordered = FALSE),
    hour  = factor(hour(datetime_utc)),
    year  = year(datetime_utc),
    season = case_when(
      month(datetime_utc) %in% c(12, 1, 2) ~ "Winter",
      month(datetime_utc) %in% c(3, 4, 5)  ~ "Spring",
      month(datetime_utc) %in% c(6, 7, 8)  ~ "Summer",
      TRUE                                  ~ "Autumn"
    ),
    season = factor(season, levels = c("Winter", "Spring", "Summer", "Autumn"))
  )

cat("Joined rows:", nrow(model_raw), "\n")
```

Computing capacity factors requires dividing observed generation (MW) by
installed capacity (MW). SMARD does not include an installed-capacity
column, so we construct a lookup from
[BNetzA / Marktstammdatenregister](https://www.marktstammdatenregister.de)
year-end published totals and interpolate linearly to daily resolution.

```{r}
#| label: installed-capacity

# ── Year-end installed capacity (MW) — BNetzA monitoring reports ─────
# Sources: BNetzA Monitoringbericht, SMARD "Installed generation capacity"
# These are approximate — replace with exact values if available.

installed_cap <- tribble(
  ~date,            ~cap_wind_onshore, ~cap_wind_offshore, ~cap_solar,
  "2019-12-31",     53200,             7700,               49100,
  "2020-12-31",     54400,             7770,               53700,
  "2021-12-31",     55900,             7790,               58700,
  "2022-12-31",     58100,             8050,               66500,
  "2023-12-31",     61000,             8390,               74100,
  "2024-12-31",     63800,             8900,               81700,
  "2025-06-30",     65500,             9200,               86000   # provisional
) |>
  mutate(date = as_date(date))

# ── Linearly interpolate to daily resolution ─────────────────────────
date_seq <- tibble(date = seq(min(installed_cap$date),
                              max(installed_cap$date),
                              by = "day"))

cap_daily <- date_seq |>
  left_join(installed_cap, by = "date") |>
  mutate(across(starts_with("cap_"),
                ~ approx(x = date[!is.na(.x)],
                          y = .x[!is.na(.x)],
                          xout = date)$y))
```

```{r}
#| label: compute-cf

model_data <- model_raw |>
  left_join(cap_daily, by = "date") |>
  mutate(
    cf_wind_onshore  = wind_onshore  / cap_wind_onshore,
    cf_wind_offshore = wind_offshore / cap_wind_offshore,
    cf_solar         = solar         / cap_solar
  ) |>
  # Solar CF is only meaningful during daylight; set nighttime to NA
  mutate(cf_solar = if_else(ssrd_wm2 > 0 | solar > 0,
                            cf_solar,
                            NA_real_))

cat("Observations with valid CF:",
    sum(!is.na(model_data$cf_wind_onshore)), "(wind on),",
    sum(!is.na(model_data$cf_wind_offshore)), "(wind off),",
    sum(!is.na(model_data$cf_solar)), "(solar)\n")
```


## 5.0 Correlation structure {#sec-correlations}

```{r}
#| label: fig-correlation
#| fig-height: 6

cor_vars <- model_data |>
  select(
    `Wind 100m`   = wind_speed_100m,
    `Wind 10m`    = wind_speed_10m,
    `Temp 2m`     = temperature_2m,
    `Solar rad`   = ssrd_wm2,
    `CF onshore`  = cf_wind_onshore,
    `CF offshore` = cf_wind_offshore,
    `CF solar`    = cf_solar
  ) |>
  drop_na()

cor_mat <- cor(cor_vars)

cor_long <- cor_mat |>
  as.data.frame() |>
  rownames_to_column("var1") |>
  pivot_longer(-var1, names_to = "var2", values_to = "r") |>
  mutate(
    var1 = factor(var1, levels = colnames(cor_mat)),
    var2 = factor(var2, levels = colnames(cor_mat))
  ) |>
  filter(as.integer(var1) >= as.integer(var2))

ggplot(cor_long, aes(x = var2, y = var1, fill = r)) +
  geom_tile(colour = "white", linewidth = 0.5) +
  geom_text(aes(label = sprintf("%.2f", r)), size = 3.2) +
  scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B",
                       midpoint = 0, limits = c(-1, 1)) +
  labs(
    title    = "Weather–generation correlation matrix",
    subtitle = glue("{zone_file}, hourly observations"),
    fill     = "Pearson r",
    caption  = "Sources: Copernicus ERA5, SMARD (BNetzA)"
  ) +
  coord_equal() +
  theme(axis.title = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1))
```


## 6.0 Scatter plots: weather → capacity factor {#sec-scatter}

### 6.1 Wind speed → wind onshore CF {#sec-wind-scatter}

```{r}
#| label: fig-wind-scatter
#| fig-height: 5

model_data |>
  filter(!is.na(cf_wind_onshore)) |>
  slice_sample(n = 15000) |>
  ggplot(aes(x = wind_speed_100m, y = cf_wind_onshore)) +
  geom_point(alpha = 0.05, colour = pal["wind_onshore"], size = 0.5) +
  geom_smooth(method = "lm",
              formula = y ~ poly(x, 3),
              colour = "black", linewidth = 0.8, se = TRUE) +
  scale_y_continuous(labels = scales::percent_format(),
                     limits = c(0, 1)) +
  labs(
    title    = "Wind speed → onshore capacity factor",
    subtitle = glue("Degree-3 polynomial fit | {zone_file} hourly"),
    x        = "ERA5 100 m wind speed (m/s)",
    y        = "Capacity factor",
    caption  = "Sources: Copernicus ERA5, SMARD (BNetzA)"
  )
```


### 6.2 Solar irradiance → solar PV CF {#sec-solar-scatter}

The marginal relationship between irradiance and solar CF appears
non-monotonic when plotted unconditionally. This is not a modelling
problem — it reflects the mixing of seasonal regimes. At the same
irradiance level (e.g. 100 W/m²), winter midday produces a high CF
(that is as good as the sun gets in January) while summer cloud cover
at the same irradiance implies suppressed output relative to clear-sky
potential. The hour and month fixed effects in the model capture this
conditioning; the faceted plots below confirm monotonic relationships
within time-of-day strata.

```{r}
#| label: fig-solar-scatter-marginal
#| fig-height: 8
#| fig-width: 12

solar_plot_df <- model_data |>
  filter(!is.na(cf_solar), ssrd_wm2 > 5)

p_marginal <- solar_plot_df |>
  slice_sample(n = 15000) |>
  ggplot(aes(x = ssrd_wm2, y = cf_solar)) +
  geom_point(alpha = 0.05, colour = pal["solar"], size = 0.5) +
  geom_smooth(method = "gam", formula = y ~ s(x),
              colour = "black", linewidth = 0.8, se = TRUE) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Marginal (unconditional)",
    x     = expression("W/m"^2),
    y     = "Capacity factor"
  )

p_faceted <- solar_plot_df |>
  mutate(hour_bin = cut(hour(datetime_utc),
                        breaks = seq(4, 22, by = 2),
                        labels = paste0(seq(4, 20, by = 2), "–",
                                        seq(6, 22, by = 2)),
                        include.lowest = TRUE)) |>
  filter(!is.na(hour_bin)) |>
  slice_sample(n = 20000) |>
  ggplot(aes(x = ssrd_wm2, y = cf_solar)) +
  geom_point(alpha = 0.06, colour = pal["solar"], size = 0.3) +
  geom_smooth(method = "gam", formula = y ~ s(x),
              colour = "black", linewidth = 0.7) +
  facet_wrap(~ hour_bin, ncol = 3) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Conditioned on time of day (UTC)",
    x     = expression("W/m"^2),
    y     = "Capacity factor"
  )

p_marginal | p_faceted +
  plot_annotation(
    title   = "Solar irradiance → PV capacity factor",
    caption = "Sources: Copernicus ERA5, SMARD (BNetzA)"
  )
```


### 6.3 Midday deep dive: season resolves the remaining non-monotonicity {#sec-solar-midday}

Even within the 10–12 UTC bin, the relationship appears non-monotonic.
Faceting by season confirms this is a **solar geometry effect**: the
same irradiance value implies very different things in winter versus
summer.

```{r}
#| label: fig-solar-midday-season
#| fig-height: 5
#| fig-width: 10

model_data |>
  filter(!is.na(cf_solar), ssrd_wm2 > 5,
         hour(datetime_utc) %in% 10:11) |>
  ggplot(aes(x = ssrd_wm2, y = cf_solar)) +
  geom_point(alpha = 0.08, colour = pal["solar"], size = 0.5) +
  geom_smooth(method = "gam", formula = y ~ s(x),
              colour = "black", linewidth = 0.8) +
  facet_wrap(~ season) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title    = "10–12 UTC: irradiance → CF by season",
    subtitle = "Within-season monotonicity confirms need for month × irradiance interactions",
    x        = expression("W/m"^2),
    y        = "Capacity factor",
    caption  = "Sources: Copernicus ERA5, SMARD (BNetzA)"
  )
```


## 7.0 Capacity factor models {#sec-models}

We fit one `tidymodels` workflow per technology using a temporal
80/20 train/test split (no shuffling — critical for time-series
validity). Wind models use polynomial wind-speed terms with month/hour
fixed effects. The solar model includes **month × irradiance
interaction terms** to capture the seasonal variation in the
irradiance–CF slope identified in the EDA above.


### 7.1 Wind onshore {#sec-model-wind-on}

```{r}
#| label: model-wind-onshore

wind_on_df <- model_data |>
  filter(!is.na(cf_wind_onshore)) |>
  select(datetime_utc, cf_wind_onshore,
         wind_speed_100m, wind_speed_10m, temperature_2m,
         month, hour)

split_wind_on <- initial_time_split(wind_on_df, prop = 0.8)
cat("Train:", nrow(training(split_wind_on)),
    "| Test:", nrow(testing(split_wind_on)), "\n")

rec_wind_on <- recipe(
  cf_wind_onshore ~ wind_speed_100m + wind_speed_10m +
                     temperature_2m + month + hour,
  data = training(split_wind_on)
) |>
  step_poly(wind_speed_100m, degree = 3) |>
  step_poly(wind_speed_10m,  degree = 2) |>
  step_dummy(all_nominal_predictors())

wf_wind_on <- workflow() |>
  add_recipe(rec_wind_on) |>
  add_model(linear_reg())

fit_wind_on <- fit(wf_wind_on, data = training(split_wind_on))

preds_wind_on <- fit_wind_on |>
  augment(testing(split_wind_on))

metrics_wind_on <- preds_wind_on |>
  metrics(truth = cf_wind_onshore, estimate = .pred)

metrics_wind_on |>
  knitr::kable(digits = 4, caption = "Wind onshore model — test set performance")
```


### 7.2 Wind offshore {#sec-model-wind-off}

```{r}
#| label: model-wind-offshore

wind_off_df <- model_data |>
  filter(!is.na(cf_wind_offshore)) |>
  select(datetime_utc, cf_wind_offshore,
         wind_speed_100m, wind_speed_10m, temperature_2m,
         month, hour)

split_wind_off <- initial_time_split(wind_off_df, prop = 0.8)

rec_wind_off <- recipe(
  cf_wind_offshore ~ wind_speed_100m + wind_speed_10m +
                      temperature_2m + month + hour,
  data = training(split_wind_off)
) |>
  step_poly(wind_speed_100m, degree = 3) |>
  step_poly(wind_speed_10m,  degree = 2) |>
  step_dummy(all_nominal_predictors())

wf_wind_off <- workflow() |>
  add_recipe(rec_wind_off) |>
  add_model(linear_reg())

fit_wind_off <- fit(wf_wind_off, data = training(split_wind_off))

preds_wind_off <- fit_wind_off |>
  augment(testing(split_wind_off))

metrics_wind_off <- preds_wind_off |>
  metrics(truth = cf_wind_offshore, estimate = .pred)

metrics_wind_off |>
  knitr::kable(digits = 4, caption = "Wind offshore model — test set performance")
```


### 7.3 Solar PV {#sec-model-solar}

The solar recipe includes month × irradiance interactions so that the
irradiance–CF slope can vary by month — physically motivated by the
seasonal change in solar geometry (sun angle, day length) identified in
@sec-solar-midday. Temperature is included to capture the ~0.4%/°C
efficiency loss above standard test conditions (25 °C).

```{r}
#| label: model-solar

solar_df <- model_data |>
  filter(!is.na(cf_solar), ssrd_wm2 > 0) |>
  select(datetime_utc, cf_solar,
         ssrd_wm2, temperature_2m,
         month, hour)

split_solar <- initial_time_split(solar_df, prop = 0.8)

rec_solar <- recipe(
  cf_solar ~ ssrd_wm2 + temperature_2m + month + hour,
  data = training(split_solar)
) |>
  step_dummy(all_nominal_predictors()) |>
  step_interact(terms = ~ ssrd_wm2:starts_with("month_"))

wf_solar <- workflow() |>
  add_recipe(rec_solar) |>
  add_model(linear_reg())

fit_solar <- fit(wf_solar, data = training(split_solar))

preds_solar <- fit_solar |>
  augment(testing(split_solar))

metrics_solar <- preds_solar |>
  metrics(truth = cf_solar, estimate = .pred)

metrics_solar |>
  knitr::kable(digits = 4, caption = "Solar PV model — test set performance")
```


## 8.0 Model diagnostics {#sec-diagnostics}

Two diagnostic panels per technology: predicted versus actual capacity
factor (a well-calibrated model hugs the 1:1 line) and the residual
distribution (centred near zero with no heavy skew indicates the model
specification is adequate).

```{r}
#| label: fig-diagnostics
#| fig-height: 10
#| fig-width: 10

diag_panels <- function(preds, truth_col, label, colour) {
  n_plot <- min(8000, nrow(preds))

  p_scatter <- preds |>
    slice_sample(n = n_plot) |>
    ggplot(aes(x = .data[[truth_col]], y = .pred)) +
    geom_point(alpha = 0.05, colour = colour, size = 0.5) +
    geom_abline(slope = 1, intercept = 0,
                linetype = "dashed", colour = "grey30") +
    scale_x_continuous(labels = scales::percent_format()) +
    scale_y_continuous(labels = scales::percent_format()) +
    coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
    labs(title = label, x = "Actual CF", y = "Predicted CF")

  p_resid <- preds |>
    mutate(resid = .pred - .data[[truth_col]]) |>
    ggplot(aes(x = resid)) +
    geom_histogram(bins = 80, fill = colour, alpha = 0.7) +
    geom_vline(xintercept = 0, linetype = "dashed", colour = "grey30") +
    labs(title = paste(label, "— residuals"),
         x = "Residual (pred − actual)", y = "Count")

  p_scatter + p_resid
}

p1 <- diag_panels(preds_wind_on,  "cf_wind_onshore",  "Wind onshore",  pal["wind_onshore"])
p2 <- diag_panels(preds_wind_off, "cf_wind_offshore", "Wind offshore", pal["wind_offshore"])
p3 <- diag_panels(preds_solar,    "cf_solar",         "Solar PV",      pal["solar"])

(p1 / p2 / p3) +
  plot_annotation(
    title   = glue("Capacity factor model diagnostics — {zone_file} test set"),
    caption = "Sources: Copernicus ERA5, SMARD (BNetzA)"
  )
```


## 9.0 Save fitted workflows {#sec-export}

The fitted `tidymodels` workflow objects are serialised to `models/` for
reuse in Stage 2 (price decomposition) and Stage 3 (risk simulation).
Each `.rds` file contains the full preprocessing recipe plus fitted
coefficients — `predict(readRDS(...), new_data)` is all downstream
stages need.

```{r}
#| label: save-models

dir.create("models", showWarnings = FALSE)

saveRDS(fit_wind_on,  glue("models/cf_wind_onshore_{zone_file}.rds"))
saveRDS(fit_wind_off, glue("models/cf_wind_offshore_{zone_file}.rds"))
saveRDS(fit_solar,    glue("models/cf_solar_{zone_file}.rds"))

cat(glue("Saved 3 workflow objects to models/ ({zone_file})"), "\n")
```


## 10.0 Key findings {#sec-findings}

```{r}
#| label: summary-table

bind_rows(
  metrics_wind_on  |> mutate(technology = "Wind onshore"),
  metrics_wind_off |> mutate(technology = "Wind offshore"),
  metrics_solar    |> mutate(technology = "Solar PV")
) |>
  select(technology, .metric, .estimate) |>
  pivot_wider(names_from = .metric, values_from = .estimate) |>
  knitr::kable(
    digits = 4,
    caption = glue("Summary of capacity factor model performance on the temporal test set ({zone_file})")
  )
```

**Interpretation.** ERA5 100 m wind speed is the dominant predictor of
wind generation capacity factors, with the cubic polynomial capturing
the characteristic ramp-up and saturation of a turbine power curve.
Solar irradiance maps onto PV capacity factors with a slope that varies
by month — an interaction effect driven by solar geometry (sun angle and
day length). The month × irradiance terms resolve the apparent
non-monotonicity visible in the unconditional scatter plot (@sec-solar-scatter),
confirming that the bimodal pattern reflects seasonal regime mixing
rather than a data quality issue. Temperature adds a modest correction
for PV cell-efficiency losses above standard test conditions (25 °C).

These fitted workflows feed directly into Stage 2, where weather-driven
generation predictions are used as features in the day-ahead price
decomposition model.
