---
title: "Price Decomposition"
subtitle: "Attributing Day-Ahead Price Variance to Renewable Variability"
format:
  html:
    code-fold: true
    code-summary: "Show code"
    toc: true
    toc-depth: 3
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| code-fold: false

library(tidyverse)
library(tidymodels)
library(arrow)
library(glue)
library(tidyquant)
library(imputeTS)
library(iml)
library(car)       # VIF
library(lmtest)    # Breusch-Pagan test
library(sandwich)  # HC-robust standard errors

theme_report <- theme_minimal(base_size = 12) +
  theme(
    plot.title    = element_text(face = "bold"),
    plot.subtitle = element_text(colour = "grey40"),
    strip.text    = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
theme_set(theme_report)

knitr::opts_chunk$set(fig.width = 10, fig.height = 6, cache.lazy = FALSE)
```

# 1.0 Introduction {#sec-intro}

This page is the analytical centrepiece of the project. We take hourly generation, load, cross-border flow, and gas price data for the DE-LU bidding zone and decompose day-ahead electricity price variance into its key drivers: renewable penetration, residual demand, interconnector flows, fuel costs, and temporal patterns. The goal is to answer a single question with precision:

> *How much of day-ahead price variance is attributable to renewable generation variability — and how does this change by season?*


# 2.0 Data {#sec-data}

## 2.1 SMARD (DE-LU)

```{r}
#| label: load-smard

smard <- read_parquet("data/processed/smard/smard_DE-LU_2020_2025.parquet")

cat(
  "SMARD rows:", nrow(smard), "\n",
  "Range:", as.character(min(smard$datetime_utc)),
  "to", as.character(max(smard$datetime_utc)), "\n",
  "Columns:", paste(names(smard), collapse = ", ")
)
```

## 2.2 Cross-Border Flows (ENTSO-E)

```{r}
#| label: load-flows

flows <- read_parquet("data/processed/entsoe/crossborder_flows_2020_2025.parquet")

cat(
  "Flow rows:", nrow(flows), "\n",
  "Pairs:", ncol(flows) - 1, "\n"
)

# ENTSO-E switched from hourly to 15-min reporting during 2021
flows |>
  mutate(year = year(datetime_utc)) |>
  count(year, name = "rows") |>
  mutate(resolution = if_else(rows > 8784 * 2, "15-min", "hourly")) |>
  print()
```

## 2.3 TTF Gas Price (Yahoo Finance)

Daily Dutch TTF front-month settlement in EUR/MWh, pulled via `tidyquant`. Saved to parquet on first run to avoid re-downloading on subsequent renders.

::: {.callout-note}
A production system would source TTF from a market data vendor (e.g. ICE, EEX). For this portfolio project, Yahoo Finance provides the same front-month TTF series at daily granularity, which is the natural resolution for a gas price proxy — settlement prices don't move intra-day.
:::

```{r}
#| label: load-ttf

ttf_path <- "data/processed/ttf/ttf_daily.parquet"

if (!file.exists(ttf_path)) {
  ttf_raw <- tq_get("TTF=F", get = "stock.prices", from = "2020-01-01")

  ttf_daily <- ttf_raw |>
    transmute(date, ttf_eur_mwh = adjusted) |>
    filter(!is.na(ttf_eur_mwh))

  dir.create(dirname(ttf_path), recursive = TRUE, showWarnings = FALSE)
  write_parquet(ttf_daily, ttf_path)
  cat("TTF downloaded and saved:", nrow(ttf_daily), "trading days\n")
} else {
  ttf_daily <- read_parquet(ttf_path)
  cat("TTF loaded from cache:", nrow(ttf_daily), "trading days\n")
}

cat(
  "Range:", as.character(min(ttf_daily$date)),
  "to", as.character(max(ttf_daily$date)), "\n",
  "TTF min:", round(min(ttf_daily$ttf_eur_mwh), 1),
  "| max:", round(max(ttf_daily$ttf_eur_mwh), 1), "EUR/MWh"
)
```


# 3.0 Exploratory Analysis — Raw Inputs {#sec-eda-raw}

Before constructing any features, we inspect the raw price series, its distributional shape, temporal structure, and relationship to gas-fired generation.

## 3.1 Price Time Series

```{r}
#| label: fig-price-ts
#| fig-cap: "DE-LU day-ahead price, hourly. LOESS trend highlights the 2021-22 energy crisis and subsequent normalisation."

smard |>
  ggplot(aes(datetime_utc, price_de_lu)) +
  geom_line(alpha = 0.15, linewidth = 0.15, colour = "grey40") +
  geom_smooth(method = "loess", span = 0.05, se = FALSE,
              colour = "#E63946", linewidth = 0.8) +
  labs(title = "DE-LU Day-Ahead Electricity Price",
       x = NULL, y = "EUR/MWh") +
  scale_y_continuous(labels = scales::comma_format())
```

## 3.2 Price Distribution by Year

```{r}
#| label: fig-price-dist
#| fig-cap: "Density of hourly prices by year. 2022 shows the heavy right tail driven by gas price pass-through."

smard |>
  mutate(year = factor(year(datetime_utc))) |>
  ggplot(aes(price_de_lu, fill = year, colour = year)) +
  geom_density(alpha = 0.25, linewidth = 0.5) +
  coord_cartesian(xlim = c(-50, 400)) +
  labs(title = "Price Distribution by Year",
       x = "EUR/MWh", y = "Density", fill = "Year", colour = "Year") +
  scale_fill_brewer(palette = "Set2") +
  scale_colour_brewer(palette = "Set2")
```

## 3.3 Negative Prices

```{r}
#| label: tbl-negative-prices
#| tbl-cap: "Negative price events by year. These reflect renewable oversupply and conventional inflexibility."

smard |>
  mutate(year = year(datetime_utc)) |>
  summarise(
    n_negative = sum(price_de_lu < 0, na.rm = TRUE),
    pct_negative = mean(price_de_lu < 0, na.rm = TRUE),
    min_price = min(price_de_lu, na.rm = TRUE),
    .by = year
  ) |>
  mutate(pct_negative = scales::percent(pct_negative, accuracy = 0.1)) |>
  knitr::kable(col.names = c("Year", "Negative hours", "% Negative", "Min EUR/MWh"),
               digits = 1)
```

```{r}
#| label: fig-negative-ts
#| fig-cap: "Negative price events coloured by magnitude. Most cluster in midday hours during high-solar periods."

smard |>
  filter(price_de_lu < 0) |>
  ggplot(aes(datetime_utc, price_de_lu, colour = price_de_lu)) +
  geom_point(size = 0.8, alpha = 0.7) +
  scale_colour_viridis_c(option = "plasma", direction = -1, name = "EUR/MWh") +
  labs(title = "Negative Price Events — DE-LU",
       x = NULL, y = "EUR/MWh") +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50")
```

::: {.callout-note title="Why Negative Prices Matter"}
Negative prices reflect the inflexibility cost of conventional generation. When renewable output exceeds demand, thermal plants that cannot ramp down quickly (lignite, nuclear, combined heat-and-power under must-run obligations) effectively pay to remain connected. For a renewable generator without a power purchase agreement, these hours represent direct revenue losses. Negative prices are preserved in the model — they are not excluded or transformed away, because they represent a structurally important feature of electricity markets that any hedging strategy must account for.
:::

## 3.4 Seasonal Heatmap (Hour x Day-of-Week)

```{r}
#| label: fig-heatmap-raw
#| fig-cap: "Average hourly price by day-of-week, hour, and season. The weekday peak-hour premium is most pronounced in winter; summer shows a midday solar dip."
#| fig-height: 8

smard |>
  mutate(
    hour   = hour(datetime_utc),
    dow    = wday(datetime_utc, label = TRUE, week_start = 1),
    season = case_when(
      month(datetime_utc) %in% c(12, 1, 2) ~ "Winter",
      month(datetime_utc) %in% 3:5         ~ "Spring",
      month(datetime_utc) %in% 6:8         ~ "Summer",
      month(datetime_utc) %in% 9:11        ~ "Autumn"
    ) |> factor(levels = c("Winter", "Spring", "Summer", "Autumn"))
  ) |>
  summarise(avg_price = mean(price_de_lu, na.rm = TRUE),
            .by = c(hour, dow, season)) |>
  ggplot(aes(hour, dow, fill = avg_price)) +
  geom_tile() +
  facet_wrap(~ season) +
  scale_fill_viridis_c(option = "inferno", name = "EUR/MWh") +
  scale_x_continuous(breaks = seq(0, 23, 3)) +
  labs(title = "Average Price Heatmap — Hour x Day-of-Week x Season",
       x = "Hour (UTC)", y = NULL)
```

## 3.5 Price vs Gas-Fired Generation

```{r}
#| label: fig-price-gas
#| fig-cap: "Higher gas dispatch is associated with higher prices — gas plants typically set the marginal price."

smard |>
  ggplot(aes(gas / 1e3, price_de_lu)) +
  geom_hex(bins = 80) +
  scale_fill_viridis_c(option = "mako", trans = "log10", name = "Count") +
  labs(title = "Price vs Gas-Fired Generation",
       x = "Gas Generation (GW)", y = "EUR/MWh") +
  coord_cartesian(ylim = c(-50, 400))
```

## 3.6 TTF Gas Price

```{r}
#| label: fig-ttf-ts
#| fig-cap: "Daily TTF front-month settlement. The 2022 spike above 300 EUR/MWh drove record electricity prices."

ttf_daily |>
  ggplot(aes(date, ttf_eur_mwh)) +
  geom_line(colour = "#457B9D", linewidth = 0.5) +
  labs(title = "TTF Natural Gas — Front-Month Settlement",
       x = NULL, y = "EUR/MWh")
```


# 4.0 Feature Engineering {#sec-features}

We construct the variables that enter the price regression. Each captures a distinct channel through which supply and demand shape the clearing price:

- **Renewable share** — wind + solar as a fraction of total load. A merit-order shifter: higher renewable share pushes expensive thermal plants off the stack.
- **Residual load** — total load minus renewable generation. The demand that must be met by dispatchable (mostly thermal) capacity.
- **Net export balance** — total physical exports minus imports across DE-LU borders. Positive = Germany is a net exporter (tightens domestic supply).
- **TTF gas price** — the marginal fuel cost for gas-fired generation, which frequently sets the day-ahead clearing price.

```{r}
#| label: feature-engineering

# --- Identify DE-LU flow columns ---
flow_names <- names(flows) |> str_subset("datetime", negate = TRUE)
export_cols <- flow_names[str_detect(flow_names, "^DE[_-]?LU")]
import_cols <- flow_names[str_detect(flow_names, "DE[_-]?LU$")]

cat("Export columns:", length(export_cols), "\n")
cat("Import columns:", length(import_cols), "\n")

# --- Compute net exports ---
# ENTSO-E flows switch from hourly to 15-minute resolution partway through
# 2021 (2022 onward is fully quarter-hourly). Aggregate to hourly means
# before joining with the hourly SMARD data.
net_exports <- flows |>
  mutate(
    total_export = rowSums(pick(all_of(export_cols)), na.rm = TRUE),
    total_import = rowSums(pick(all_of(import_cols)), na.rm = TRUE),
    net_export_mw = total_export - total_import,
    datetime_utc  = floor_date(datetime_utc, unit = "hour")
  ) |>
  summarise(net_export_mw = mean(net_export_mw), .by = datetime_utc)

cat("Net exports after hourly aggregation:", nrow(net_exports), "rows\n")

# --- Build feature table ---
price_df <- smard |>
  mutate(
    renewable_gen   = wind_onshore + wind_offshore + solar,
    renewable_share = renewable_gen / total_load,
    residual_load   = total_load - renewable_gen,
    date   = as_date(datetime_utc),
    month  = month(datetime_utc),
    hour   = hour(datetime_utc),
    dow    = wday(datetime_utc, label = TRUE, week_start = 1),
    season = case_when(
      month %in% c(12, 1, 2) ~ "Winter",
      month %in% 3:5         ~ "Spring",
      month %in% 6:8         ~ "Summer",
      month %in% 9:11        ~ "Autumn"
    ) |> factor(levels = c("Winter", "Spring", "Summer", "Autumn"))
  ) |>
  # Join net exports
  left_join(net_exports, by = "datetime_utc") |>
  # Join TTF (daily step series — forward-fill weekends/holidays)
  left_join(ttf_daily, by = "date") |>
  arrange(datetime_utc) |>
  fill(ttf_eur_mwh, .direction = "down")

# --- Summary ---
cat("\nFeature table:", nrow(price_df), "rows\n")
cat("NAs after join:\n")
price_df |>
  summarise(across(c(renewable_share, residual_load, net_export_mw, ttf_eur_mwh),
                   ~ sum(is.na(.)))) |>
  glimpse()
```

```{r}
#| label: impute-features
#| code-fold: false

# Impute remaining feature gaps rather than dropping rows.
# TTF: back-fill leading NAs (before first trading day); the forward-fill
#   above already handles weekends/holidays.
# Net exports, renewable share, residual load: Kalman smoothing on a
#   structural time series model (imputeTS) — respects the autocorrelation
#   structure of hourly energy data better than linear interpolation.
# Nuclear is NOT imputed — post-April-2023 NAs/zeros reflect the actual
#   phase-out of German nuclear generation.

price_df <- price_df |>
  arrange(datetime_utc) |>
  fill(ttf_eur_mwh, .direction = "downup") |>
  mutate(across(
    c(renewable_share, residual_load, net_export_mw),
    ~ imputeTS::na_kalman(., model = "StructTS", smooth = TRUE)
  ))

remaining_na <- price_df |>
  summarise(across(c(price_de_lu, renewable_share, residual_load,
                     net_export_mw, ttf_eur_mwh),
                   ~ sum(is.na(.))))

cat("Rows retained:", nrow(price_df), "\n")
cat("Remaining NAs after imputation:\n")
glimpse(remaining_na)

# Drop any rows where price itself is missing (target variable — can't impute)
price_df <- price_df |> drop_na(price_de_lu)
cat("Final rows (after dropping missing prices):", nrow(price_df))
```


# 5.0 Exploratory Analysis — Constructed Features {#sec-eda-features}

Now that the features are built, we inspect their relationships with price before modelling.

## 5.1 Price vs Renewable Share by Season

```{r}
#| label: fig-price-renshare
#| fig-cap: "Negative relationship in all seasons: higher renewable share depresses prices. The slope is steepest in winter, when displaced thermal capacity is most expensive."

price_df |>
  ggplot(aes(renewable_share, price_de_lu)) +
  geom_hex(bins = 60) +
  geom_smooth(method = "lm", colour = "#E63946", se = FALSE, linewidth = 0.7) +
  facet_wrap(~ season) +
  scale_fill_viridis_c(option = "mako", trans = "log10", name = "Count") +
  coord_cartesian(ylim = c(-50, 300)) +
  scale_x_continuous(labels = scales::percent_format()) +
  labs(title = "Price vs Renewable Share by Season",
       x = "Renewable Share (wind + solar / load)", y = "EUR/MWh")
```

## 5.2 Price vs Residual Load

```{r}
#| label: fig-price-resload
#| fig-cap: "Price increases non-linearly with residual load — the merit order curve steepens as thermal capacity tightens."

price_df |>
  ggplot(aes(residual_load / 1e3, price_de_lu)) +
  geom_hex(bins = 80) +
  geom_smooth(method = "gam", formula = y ~ s(x), colour = "#E63946",
              se = FALSE, linewidth = 0.7) +
  scale_fill_viridis_c(option = "mako", trans = "log10", name = "Count") +
  coord_cartesian(ylim = c(-50, 400)) +
  labs(title = "Price vs Residual Load",
       x = "Residual Load (GW)", y = "EUR/MWh")
```

## 5.3 Price vs Net Exports

```{r}
#| label: fig-price-exports
#| fig-cap: "When Germany is a net exporter (positive values), domestic supply is tighter and prices tend to be higher."

price_df |>
  ggplot(aes(net_export_mw / 1e3, price_de_lu)) +
  geom_hex(bins = 80) +
  geom_smooth(method = "gam", formula = y ~ s(x), colour = "#E63946",
              se = FALSE, linewidth = 0.7) +
  scale_fill_viridis_c(option = "mako", trans = "log10", name = "Count") +
  coord_cartesian(ylim = c(-50, 400)) +
  labs(title = "Price vs Net Export Position",
       x = "Net Exports (GW) — positive = exporting", y = "EUR/MWh")
```

## 5.4 Feature Correlation Matrix

```{r}
#| label: fig-corr-matrix
#| fig-cap: "Correlation among model features. Renewable share and residual load are strongly negatively correlated by construction — this is expected and handled by the lasso."

cor_vars <- price_df |>
  select(price_de_lu, renewable_share, residual_load, net_export_mw, ttf_eur_mwh)

cor_mat <- cor(cor_vars, use = "complete.obs")

cor_mat |>
  as_tibble(rownames = "var1") |>
  pivot_longer(-var1, names_to = "var2", values_to = "r") |>
  mutate(
    var1 = str_replace_all(var1, "_", " "),
    var2 = str_replace_all(var2, "_", " ")
  ) |>
  ggplot(aes(var1, var2, fill = r)) +
  geom_tile() +
  geom_text(aes(label = round(r, 2)), size = 3.5) +
  scale_fill_gradient2(low = "#457B9D", mid = "white", high = "#E63946",
                       midpoint = 0, limits = c(-1, 1), name = "r") +
  labs(title = "Feature Correlation Matrix", x = NULL, y = NULL) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


# 6.0 Train / Test Split {#sec-split}

We use an 80/20 temporal split — the model is trained on the earlier period and evaluated on the most recent data. This respects time ordering and prevents data leakage.

```{r}
#| label: train-test-split
#| code-fold: false

split_date <- quantile(as.numeric(price_df$datetime_utc), 0.8) |>
  as.POSIXct(origin = "1970-01-01", tz = "UTC")

train_df <- price_df |> filter(datetime_utc < split_date)
test_df  <- price_df |> filter(datetime_utc >= split_date)

cat(
  "Train:", nrow(train_df), "rows |",
  as.character(min(train_df$datetime_utc)), "to",
  as.character(max(train_df$datetime_utc)), "\n",
  "Test: ", nrow(test_df), "rows |",
  as.character(min(test_df$datetime_utc)), "to",
  as.character(max(test_df$datetime_utc))
)
```


# 7.0 Pre-Modelling Diagnostics {#sec-diagnostics}

Before fitting the price model, we check the standard regression assumptions against a preliminary OLS fit to motivate modelling choices.

```{r}
#| label: diagnostics-ols-prelim

diag_fml <- price_de_lu ~ renewable_share + residual_load + net_export_mw +
  ttf_eur_mwh + season + dow + hour
diag_fit <- lm(diag_fml, data = train_df)
```

## 7.1 Variance Inflation Factors (VIF)

```{r}
#| label: tbl-vif
#| tbl-cap: "Generalised VIF for each predictor. Values above 5-10 indicate collinearity concerns that motivate regularisation."

car::vif(diag_fit) |>
  as.data.frame() |>
  rownames_to_column("term") |>
  as_tibble() |>
  rename(gvif = GVIF, df = Df, gvif_adj = `GVIF^(1/(2*Df))`) |>
  mutate(across(where(is.numeric), ~ round(., 2))) |>
  arrange(desc(gvif_adj)) |>
  knitr::kable()
```

## 7.2 Residuals vs Fitted

```{r}
#| label: fig-resid-fitted
#| fig-cap: "Residual variance increases with fitted value — clear heteroskedasticity. This is a structural feature of electricity markets: gas-fired marginal pricing produces multiplicative price responses."

tibble(fitted = fitted(diag_fit), residual = residuals(diag_fit)) |>
  ggplot(aes(fitted, residual)) +
  geom_hex(bins = 80) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_smooth(method = "loess", se = FALSE, colour = "#E63946", linewidth = 0.7) +
  scale_fill_viridis_c(option = "mako", trans = "log10", name = "Count") +
  labs(title = "Residuals vs Fitted Values (Preliminary OLS)",
       x = "Fitted (EUR/MWh)", y = "Residual (EUR/MWh)")
```

## 7.3 Q-Q Plot

```{r}
#| label: fig-qq
#| fig-cap: "Heavy tails relative to normality — extreme price events are more frequent than a Gaussian model implies. OLS consistency does not require normality, but this reinforces the need for robust inference."

tibble(residual = residuals(diag_fit)) |>
  ggplot(aes(sample = residual)) +
  stat_qq(alpha = 0.05, size = 0.5, colour = "grey40") +
  stat_qq_line(colour = "#E63946", linewidth = 0.7) +
  labs(title = "Q-Q Plot — OLS Residuals",
       x = "Theoretical Quantiles", y = "Sample Quantiles (EUR/MWh)")
```

## 7.4 Breusch-Pagan Test

```{r}
#| label: breusch-pagan

bp_test <- lmtest::bptest(diag_fit)
cat(
  "Breusch-Pagan statistic:", round(bp_test$statistic, 1), "\n",
  "p-value:", format.pval(bp_test$p.value, digits = 3), "\n\n"
)
if (bp_test$p.value < 0.05) {
  cat("Result: heteroskedasticity detected (p < 0.05).\n",
      "This is expected for electricity prices and handled with HC-robust standard errors.\n")
}
```

::: {.callout-note title="Diagnostic Summary"}
The diagnostics reveal two features typical of electricity price data:

1. **Multicollinearity** between renewable share and residual load (mechanically linked via total load). A **lasso** penalty addresses this by zeroing out redundant terms.
2. **Heteroskedasticity** — residual variance scales with price level. This is a structural feature of electricity markets, not a model failure. **HC-robust standard errors** (sandwich estimator) ensure valid inference. The model operates in raw EUR/MWh for full interpretability.

These findings motivate the modelling choices in the next section.
:::


# 8.0 Price Regression {#sec-regression}

We fit three models and compare them via rolling-window cross-validation:

1. **Lasso with interactions** — interpretable, coefficient-based. The L1 penalty produces a sparse model by zeroing out redundant terms. All numeric predictors are normalised so coefficients read as EUR/MWh per 1-SD change.
2. **Boosted tree (XGBoost, mean)** — flexible, non-parametric benchmark. Captures non-linearities and higher-order interactions the lasso cannot.
3. **Quantile XGBoost (q5, q50, q95)** — targets specific percentiles, producing a 90% prediction interval that captures tail events the mean models miss. Feeds directly into the Stage 3 VaR/CVaR framework.

The lasso is the primary model for the variance decomposition; the boosted trees serve as performance benchmarks and provide distributional predictions for risk analysis.

## 8.1 Feature Recipes

```{r}
#| label: recipes

# --- Lasso recipe: cyclical encoding, interactions, normalisation ---
lasso_recipe <- recipe(
  price_de_lu ~ renewable_share + residual_load + net_export_mw +
    ttf_eur_mwh + season + dow + hour,
  data = train_df
) |>
  step_mutate(
    hour_sin_24 = sin(2 * pi * hour / 24),
    hour_cos_24 = cos(2 * pi * hour / 24),
    hour_sin_12 = sin(4 * pi * hour / 24),
    hour_cos_12 = cos(4 * pi * hour / 24)
  ) |>
  step_rm(hour) |>
  step_dummy(season, dow) |>
  step_interact(terms = ~ renewable_share:starts_with("season_")) |>
  step_interact(terms = ~ starts_with("hour_"):starts_with("season_")) |>
  step_normalize(all_numeric_predictors())

# --- XGBoost recipe: trees handle non-linearities natively ---
xgb_recipe <- recipe(
  price_de_lu ~ renewable_share + residual_load + net_export_mw +
    ttf_eur_mwh + season + dow + hour,
  data = train_df
) |>
  step_dummy(season, dow)
```

## 8.2 Model Specifications

```{r}
#| label: model-specs

# Pure lasso: mixture = 1 (L1 only), only tune penalty
lasso_spec <- linear_reg(
  penalty = tune(),
  mixture = 1
) |>
  set_engine("glmnet") |>
  set_mode("regression")

xgb_spec <- boost_tree(
  trees      = 500,
  tree_depth = 6,
  learn_rate = 0.05,
  min_n      = 20
) |>
  set_engine("xgboost") |>
  set_mode("regression")

# Quantile XGBoost for tail risk
xgb_q05_spec <- boost_tree(trees = 500, tree_depth = 6, learn_rate = 0.05, min_n = 20) |>
  set_engine("xgboost", objective = "reg:quantileerror", quantile_alpha = 0.05) |>
  set_mode("regression")

xgb_q50_spec <- boost_tree(trees = 500, tree_depth = 6, learn_rate = 0.05, min_n = 20) |>
  set_engine("xgboost", objective = "reg:quantileerror", quantile_alpha = 0.50) |>
  set_mode("regression")

xgb_q95_spec <- boost_tree(trees = 500, tree_depth = 6, learn_rate = 0.05, min_n = 20) |>
  set_engine("xgboost", objective = "reg:quantileerror", quantile_alpha = 0.95) |>
  set_mode("regression")

lasso_wf <- workflow() |> add_recipe(lasso_recipe) |> add_model(lasso_spec)
xgb_wf   <- workflow() |> add_recipe(xgb_recipe)  |> add_model(xgb_spec)

xgb_q05_wf <- workflow() |> add_recipe(xgb_recipe) |> add_model(xgb_q05_spec)
xgb_q50_wf <- workflow() |> add_recipe(xgb_recipe) |> add_model(xgb_q50_spec)
xgb_q95_wf <- workflow() |> add_recipe(xgb_recipe) |> add_model(xgb_q95_spec)
```

## 8.3 Rolling-Window Cross-Validation

```{r}
#| label: rolling-cv

rolling_splits <- rolling_origin(
  train_df,
  initial  = 365 * 24,
  assess   = 30 * 24,
  skip     = 30 * 24,
  cumulative = FALSE
)

cat("Rolling CV folds:", nrow(rolling_splits))
```

```{r}
#| label: cv-fit
#| cache: true

cv_metrics <- metric_set(rmse, mae, rsq)

# --- Tune lasso penalty ---
lasso_grid <- grid_regular(penalty(range = c(-2, 1)), levels = 20)

lasso_tune <- tune_grid(
  lasso_wf,
  resamples = rolling_splits,
  grid      = lasso_grid,
  metrics   = cv_metrics
)

# --- XGBoost (mean): no tuning ---
xgb_cv <- fit_resamples(xgb_wf, resamples = rolling_splits, metrics = cv_metrics)

# --- Select best lasso penalty ---
best_lasso <- select_best(lasso_tune, metric = "rmse")
cat("\nBest lasso penalty:", best_lasso$penalty, "\n")

lasso_wf_final <- finalize_workflow(lasso_wf, best_lasso)
```

```{r}
#| label: tbl-cv-results
#| tbl-cap: "Rolling CV performance (EUR/MWh)."

bind_rows(
  collect_metrics(lasso_tune) |>
    semi_join(best_lasso, by = "penalty") |>
    mutate(model = "Lasso"),
  collect_metrics(xgb_cv) |> mutate(model = "XGBoost")
) |>
  select(model, .metric, mean, std_err) |>
  pivot_wider(names_from = .metric, values_from = c(mean, std_err)) |>
  knitr::kable(digits = 2)
```

## 8.4 Fit on Full Training Set & Test Evaluation

```{r}
#| label: final-fit
#| cache: true

lasso_fit <- fit(lasso_wf_final, data = train_df)
xgb_fit   <- fit(xgb_wf, data = train_df)

# Quantile XGBoost fits
xgb_q05_fit <- fit(xgb_q05_wf, data = train_df)
xgb_q50_fit <- fit(xgb_q50_wf, data = train_df)
xgb_q95_fit <- fit(xgb_q95_wf, data = train_df)

# Test-set predictions (all EUR/MWh)
test_preds <- test_df |>
  select(datetime_utc, price_de_lu, season) |>
  mutate(
    pred_lasso = predict(lasso_fit, new_data = test_df)$.pred,
    pred_xgb   = predict(xgb_fit, new_data = test_df)$.pred,
    pred_q05   = predict(xgb_q05_fit, new_data = test_df)$.pred,
    pred_q50   = predict(xgb_q50_fit, new_data = test_df)$.pred,
    pred_q95   = predict(xgb_q95_fit, new_data = test_df)$.pred
  )

bind_rows(
  test_preds |> metrics(truth = price_de_lu, estimate = pred_lasso) |> mutate(model = "Lasso"),
  test_preds |> metrics(truth = price_de_lu, estimate = pred_xgb) |> mutate(model = "XGBoost (mean)"),
  test_preds |> metrics(truth = price_de_lu, estimate = pred_q50) |> mutate(model = "XGBoost (median)")
) |>
  select(model, .metric, .estimate) |>
  pivot_wider(names_from = .metric, values_from = .estimate) |>
  knitr::kable(digits = 2, caption = "Test-Set Performance (EUR/MWh)")
```

## 8.5 Actual vs Predicted

```{r}
#| label: fig-actual-pred
#| fig-cap: "Actual vs predicted prices on the test set (EUR/MWh)."

test_preds |>
  pivot_longer(cols = c(pred_lasso, pred_xgb), names_to = "model",
               values_to = "predicted", names_prefix = "pred_") |>
  mutate(model = case_match(model, "lasso" ~ "Lasso", "xgb" ~ "XGBoost (mean)")) |>
  ggplot(aes(predicted, price_de_lu)) +
  geom_hex(bins = 60) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "#E63946") +
  facet_wrap(~ model) +
  scale_fill_viridis_c(option = "mako", trans = "log10", name = "Count") +
  coord_cartesian(xlim = c(-50, 300), ylim = c(-50, 300)) +
  labs(title = "Actual vs Predicted — Test Set (EUR/MWh)",
       x = "Predicted (EUR/MWh)", y = "Actual (EUR/MWh)")
```

## 8.5.1 Quantile Prediction Intervals

The mean-regression models compress predictions toward the centre. The quantile XGBoost models directly target the 5th and 95th percentiles, producing a 90% prediction interval that captures tail events.

```{r}
#| label: fig-quantile-interval
#| fig-cap: "Quantile XGBoost prediction intervals (90%) on the test set. The q5-q95 band brackets extreme price events that the mean models miss."

set.seed(42)
interval_sample <- test_preds |>
  arrange(pred_q50) |>
  mutate(idx = row_number()) |>
  slice_sample(n = 3000) |>
  arrange(idx)

interval_sample |>
  ggplot(aes(x = idx)) +
  geom_ribbon(aes(ymin = pred_q05, ymax = pred_q95), fill = "#457B9D", alpha = 0.3) +
  geom_line(aes(y = pred_q50), colour = "#457B9D", linewidth = 0.4) +
  geom_point(aes(y = price_de_lu), size = 0.3, alpha = 0.4, colour = "grey30") +
  labs(
    title = "90% Prediction Interval — Quantile XGBoost (EUR/MWh)",
    subtitle = "Blue band = q5-q95; line = median; points = actual prices",
    x = "Observations (sorted by predicted median)", y = "EUR/MWh"
  )
```

```{r}
#| label: tbl-quantile-coverage
#| tbl-cap: "Empirical coverage of the 90% prediction interval by season. Ideal coverage is 90%."

test_preds |>
  mutate(covered = price_de_lu >= pred_q05 & price_de_lu <= pred_q95) |>
  summarise(n_obs = n(), coverage = mean(covered), .by = season) |>
  add_row(
    season = "Overall", n_obs = nrow(test_preds),
    coverage = mean(test_preds$price_de_lu >= test_preds$pred_q05 &
                    test_preds$price_de_lu <= test_preds$pred_q95)
  ) |>
  mutate(coverage = scales::percent(coverage, accuracy = 0.1)) |>
  knitr::kable()
```

## 8.6 Residual Time Series

```{r}
#| label: fig-residuals-ts
#| fig-cap: "Daily mean lasso residuals (EUR/MWh) across the full dataset. The ribbon shows plus/minus 1 SD. The vertical line marks the train/test boundary."

full_resid <- price_df |>
  mutate(
    pred     = predict(lasso_fit, new_data = price_df)$.pred,
    residual = price_de_lu - pred,
    date     = as_date(datetime_utc),
    split    = if_else(datetime_utc < split_date, "Train", "Test")
  ) |>
  summarise(
    mean_resid = mean(residual, na.rm = TRUE),
    sd_resid   = sd(residual, na.rm = TRUE),
    split      = first(split),
    .by = date
  )

full_resid |>
  ggplot(aes(date, mean_resid)) +
  geom_ribbon(aes(ymin = mean_resid - sd_resid,
                  ymax = mean_resid + sd_resid),
              fill = "grey80", alpha = 0.5) +
  geom_line(colour = "#457B9D", linewidth = 0.4) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_vline(xintercept = as_date(split_date), linetype = "dotted",
             colour = "#E63946", linewidth = 0.5) +
  annotate("text", x = as_date(split_date),
           y = max(full_resid$mean_resid + full_resid$sd_resid, na.rm = TRUE) * 0.9,
           label = "Train / Test split", hjust = -0.05, colour = "#E63946", size = 3.5) +
  labs(title = "Daily Mean Residual — Lasso (Full Dataset)",
       subtitle = "Ribbon = +/- 1 SD of hourly residuals within each day",
       x = NULL, y = "Residual (EUR/MWh)")
```

## 8.7 Lasso Coefficients

The lasso zeroes out redundant terms, producing a sparse coefficient set. All numeric predictors are normalised, so each coefficient represents the EUR/MWh price change per 1-SD shift in that predictor — effect sizes are directly comparable across features. Only the top 20 non-zero coefficients are shown.

```{r}
#| label: tbl-lasso-coefs
#| tbl-cap: "Top 20 non-zero lasso coefficients (normalised scale, EUR/MWh per 1-SD change). Sorted by absolute magnitude."

lasso_engine <- extract_fit_engine(lasso_fit)

lasso_coefs <- coef(lasso_engine, s = best_lasso$penalty) |>
  as.matrix() |>
  as.data.frame() |>
  rownames_to_column("term") |>
  rename(estimate = 2) |>
  filter(abs(estimate) > 0, term != "(Intercept)") |>
  mutate(
    term = str_replace_all(term, "_", " "),
    estimate = round(estimate, 2)
  ) |>
  arrange(desc(abs(estimate)))

cat("Total non-zero coefficients:", nrow(lasso_coefs), "\n\n")

lasso_coefs |>
  slice_head(n = 20) |>
  knitr::kable()
```

## 8.8 OLS with HC-Robust Standard Errors

For valid statistical inference under heteroskedasticity, we refit OLS using **only the predictors selected by the lasso** (non-zero coefficients) and report HC3-robust standard errors. This two-stage approach — lasso for feature selection, OLS for inference — is a standard econometric pipeline.

```{r}
#| label: tbl-robust-coefs
#| tbl-cap: "OLS coefficients with HC3-robust standard errors on the lasso-selected features (EUR/MWh per 1-SD change)."

lasso_prep <- lasso_recipe |> prep(training = train_df)
lasso_baked_train <- bake(lasso_prep, new_data = train_df)

# Extract lasso-selected terms (non-zero at best penalty)
selected_terms <- coef(lasso_engine, s = best_lasso$penalty) |>
  as.matrix() |>
  as.data.frame() |>
  rownames_to_column("term") |>
  rename(estimate = 2) |>
  filter(abs(estimate) > 0, term != "(Intercept)") |>
  pull(term)

cat("Lasso selected", length(selected_terms), "of",
    ncol(lasso_baked_train) - 1, "candidate features\n\n")

# Refit OLS on selected features only
ols_data <- lasso_baked_train |> select(price_de_lu, all_of(selected_terms))
ols_robust_fit <- lm(price_de_lu ~ ., data = ols_data)

# HC3-robust inference
robust_coefs <- lmtest::coeftest(ols_robust_fit, vcov. = sandwich::vcovHC(ols_robust_fit, type = "HC3"))

robust_coefs |>
  broom::tidy() |>
  filter(term != "(Intercept)") |>
  mutate(
    term = str_replace_all(term, "_", " "),
    across(c(estimate, std.error, statistic), ~ round(., 2)),
    p.value = scales::pvalue(p.value)
  ) |>
  arrange(desc(abs(estimate))) |>
  slice_head(n = 20) |>
  knitr::kable()
```


# 9.0 Variance Decomposition {#sec-decomposition}

This is the centrepiece exhibit. We quantify how much of day-ahead price variance each driver explains, and show how the attribution shifts by season.

The decomposition is presented in two panels. First, the **full sample (2020-2025)** provides context: the 2021-22 gas crisis dominates overall variance, with TTF swamping all other drivers. Second, the **post-crisis subset (2023-2025)** isolates the stable gas-price regime where renewable variability, residual load, and temporal patterns emerge as the primary merit-order drivers. This second panel reflects the market environment a hedging desk faces today.

## 9.1 Full Sample — Partial R-squared by Season

```{r}
#| label: partial-r2-full
#| cache: true

# Helper function for partial R-squared decomposition
compute_partial_r2 <- function(df, response = "price_de_lu") {

  df <- df |>
    mutate(
      hour_sin_24 = sin(2 * pi * hour / 24),
      hour_cos_24 = cos(2 * pi * hour / 24),
      hour_sin_12 = sin(4 * pi * hour / 24),
      hour_cos_12 = cos(4 * pi * hour / 24)
    )

  full_fml <- reformulate(
    c("renewable_share", "residual_load", "net_export_mw", "ttf_eur_mwh",
      "dow", "hour_sin_24", "hour_cos_24", "hour_sin_12", "hour_cos_12"),
    response = response
  )

  feature_groups <- list(
    "Renewable share" = "renewable_share",
    "Residual load"   = "residual_load",
    "Net exports"     = "net_export_mw",
    "Gas price (TTF)" = "ttf_eur_mwh",
    "Temporal (hour + dow)" = c("dow", "hour_sin_24", "hour_cos_24",
                                 "hour_sin_12", "hour_cos_12")
  )

  map_dfr(levels(df$season), function(s) {
    df_s <- df |> filter(season == s)
    full_fit <- lm(full_fml, data = df_s)
    full_r2  <- summary(full_fit)$r.squared

    map_dfr(names(feature_groups), function(g) {
      drop_terms <- feature_groups[[g]]
      keep_terms <- setdiff(attr(terms(full_fml), "term.labels"), drop_terms)
      reduced_fml <- reformulate(keep_terms, response = response)
      reduced_r2  <- summary(lm(reduced_fml, data = df_s))$r.squared
      tibble(season = s, component = g, partial_r2 = full_r2 - reduced_r2)
    }) |>
      add_row(season = s, component = "Full model R-sq", partial_r2 = full_r2)
  })
}

partial_r2_full <- compute_partial_r2(price_df)
```

```{r}
#| label: fig-partial-r2-full
#| fig-cap: "Full sample (2020-2025): TTF gas price dominates variance in all seasons. The 2021-22 energy crisis was the single largest source of price variance in the dataset."
#| fig-height: 7

partial_r2_full |>
  filter(component != "Full model R-sq") |>
  mutate(
    season    = factor(season, levels = c("Winter", "Spring", "Summer", "Autumn")),
    component = fct_reorder(component, partial_r2, .fun = mean, .desc = TRUE)
  ) |>
  ggplot(aes(component, partial_r2, fill = component)) +
  geom_col(width = 0.7) +
  facet_wrap(~ season) +
  coord_flip() +
  scale_fill_brewer(palette = "Set2", guide = "none") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    title    = "Variance Attribution — Full Sample (2020-2025)",
    subtitle = "Drop in model R-squared when each feature group is excluded",
    x = NULL, y = "Partial R-squared"
  )
```

## 9.2 Post-Crisis — Partial R-squared by Season

With TTF stabilised in the 30-55 EUR/MWh range since mid-2022, the merit-order drivers that matter for forward-looking hedging become visible.

```{r}
#| label: partial-r2-post
#| cache: true

price_post <- price_df |> filter(datetime_utc >= as.POSIXct("2023-01-01", tz = "UTC"))
cat("Post-crisis rows:", nrow(price_post), "\n")

partial_r2_post <- compute_partial_r2(price_post)
```

```{r}
#| label: fig-partial-r2-post
#| fig-cap: "Post-crisis (2023-2025): with gas prices stabilised, renewable share and residual load emerge as the primary price drivers. The seasonal pattern is clear — renewables dominate summer, residual load dominates shoulder seasons."
#| fig-height: 7

partial_r2_post |>
  filter(component != "Full model R-sq") |>
  mutate(
    season    = factor(season, levels = c("Winter", "Spring", "Summer", "Autumn")),
    component = fct_reorder(component, partial_r2, .fun = mean, .desc = TRUE)
  ) |>
  ggplot(aes(component, partial_r2, fill = component)) +
  geom_col(width = 0.7) +
  facet_wrap(~ season) +
  coord_flip() +
  scale_fill_brewer(palette = "Set2", guide = "none") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    title    = "Variance Attribution — Post-Crisis (2023-2025)",
    subtitle = "Drop in model R-squared when each feature group is excluded",
    x = NULL, y = "Partial R-squared"
  )
```

```{r}
#| label: tbl-partial-r2-both
#| tbl-cap: "Partial R-squared comparison: full sample vs post-crisis, by season."

bind_rows(
  partial_r2_full |> mutate(period = "Full (2020-2025)"),
  partial_r2_post |> mutate(period = "Post-crisis (2023-2025)")
) |>
  filter(component != "Full model R-sq") |>
  mutate(partial_r2 = round(partial_r2, 4)) |>
  pivot_wider(names_from = season, values_from = partial_r2) |>
  arrange(period, desc(Winter)) |>
  knitr::kable()
```

## 9.3 Permutation Importance (Robustness Check)

As a robustness check, we compute permutation-based feature importance using the `iml` package. This approach is model-agnostic and doesn't depend on the sequential ordering of terms.

::: {.callout-note title="Partial R-squared vs Permutation Importance"}
Partial R-squared uses unpenalised OLS (fitted per season) for clean additive decomposition — it answers "how much variance is *associated with* each driver?" The permutation importance uses the fitted lasso — it answers "how much does *this specific model* rely on each feature?" If the two methods agree, that is strong evidence. If they diverge on a specific feature, the divergence flags collinearity where the lasso made an arbitrary selection between correlated features.
:::

```{r}
#| label: shapley-importance
#| cache: true

lasso_baked_test <- lasso_recipe |>
  prep(training = train_df) |>
  bake(new_data = test_df)

X_test <- lasso_baked_test |> select(-price_de_lu)
y_test <- lasso_baked_test$price_de_lu

lasso_engine_perm <- extract_fit_engine(lasso_fit)
lasso_predict <- function(model, newdata) {
  predict(model, as.matrix(newdata), s = best_lasso$penalty)[, 1]
}

predictor_obj <- Predictor$new(
  model = lasso_engine_perm,
  data  = X_test,
  y     = y_test,
  predict.function = lasso_predict
)

set.seed(42)
feat_imp <- FeatureImp$new(predictor_obj, loss = "mse", n.repetitions = 10)
```

```{r}
#| label: fig-shapley
#| fig-cap: "Permutation importance (ratio of permuted MSE to baseline MSE). Values above 1 indicate the feature contributes to model accuracy."

feat_imp$results |>
  as_tibble() |>
  mutate(feature = str_replace_all(feature, "_", " ") |> fct_reorder(importance)) |>
  ggplot(aes(importance, feature)) +
  geom_col(fill = "#457B9D", width = 0.6) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "grey50") +
  labs(
    title = "Permutation Feature Importance (Lasso on Test Set)",
    subtitle = "Ratio of permuted MSE to baseline MSE — higher = more important",
    x = "Importance (MSE ratio)", y = NULL
  )
```


# 10.0 Key Findings {#sec-findings}

::: {.callout-important title="Summary"}
1. **Fuel cost shocks dominated European electricity price variance during 2020-2022.** The TTF gas price alone explains 50-80% of seasonal price variance in the full sample — the energy crisis was the single largest source of price risk.
2. **In the post-crisis market (2023-2025), renewable variability and residual load emerge as the primary drivers.** With gas prices stabilised, the merit-order effects that shape day-ahead pricing become visible.
3. **Renewable share dominates summer price variance** in the post-crisis period, when solar generation swings prices between midday troughs and evening peaks.
4. **Residual load is most important in shoulder seasons** (spring, autumn), when neither renewable nor gas effects dominate.
5. **Net exports have a measurable but secondary effect**, consistent with Germany's central position in the CWE interconnected market.
6. **Temporal patterns (hour, day-of-week) capture structural demand cycles** — contributing consistently across all seasons.
7. **Heteroskedasticity is a structural feature of electricity markets**, not a model failure. All results are reported in raw EUR/MWh with HC-robust standard errors for valid inference.
8. **Negative prices are preserved in the model** — they reflect conventional inflexibility during renewable oversupply and are economically important for hedging.
9. **Quantile XGBoost provides distributional predictions** for risk analysis, producing a 90% prediction interval that brackets extreme price events. These feed directly into the Stage 3 VaR/CVaR framework.
:::

These findings inform the risk quantification in the next section: the seasonal attribution structure means that a hedging strategy must account for *which* risk factor dominates in the delivery period. In today's market, a summer forward position is primarily a renewable variability bet; shoulder-season contracts depend on residual load dynamics; and the ever-present tail risk of another fuel cost shock requires monitoring through the quantile framework.


::: {.callout-note title="Data Attribution"}
SMARD data: Bundesnetzagentur | SMARD.de, CC BY 4.0. ENTSO-E Transparency Platform data used per ENTSO-E terms of service. TTF gas price data sourced from Yahoo Finance (ICE Dutch TTF front-month futures). Copernicus ERA5 reanalysis: Hersbach et al. (2020), doi:10.24381/cds.adbb2d47.
:::
